---
title: "GExPipe: A Pipeline for Gene Expression Analysis"
author:
  - name: Safa Rafique, Dr. Naeem Mahmood Ashraf, Prof. Dr. Farooq Sabir
    affiliation: University of the Punjab

date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to GExPipe}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`GExPipe` is a Shiny-first workflow for bulk RNA-seq and microarray
analysis. The package was created to provide a single, guided pipeline
that connects data retrieval, quality control, normalization,
differential expression, network analysis, and downstream interpretation
without requiring users to manually stitch many separate tools together.

Within Bioconductor, `GExPipe` builds on established infrastructure such
as `r BiocStyle::Biocpkg("GEOquery")`, `r BiocStyle::Biocpkg("limma")`,
`r BiocStyle::Biocpkg("DESeq2")`, `r BiocStyle::Biocpkg("edgeR")`,
`r BiocStyle::Biocpkg("WGCNA")`, and
`r BiocStyle::Biocpkg("clusterProfiler")`. For visualization it uses
`r BiocStyle::CRANpkg("ggplot2")` and
`r BiocStyle::CRANpkg("pheatmap")`. Compared with piecemeal scripts, the
app emphasizes usability and reproducibility through a fixed
step-by-step interface and integrated summaries.

# Installation

Install `GExPipe` with `BiocManager` once the package has been accepted
on Bioconductor:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# dependencies = TRUE installs all Imports + Suggests (54 packages total)
# so every app feature works out of the box.
BiocManager::install("GExPipe", dependencies = TRUE)
```

Until then, install directly from GitHub:

> **Network requirement:** Installing from GitHub requires an internet
> connection that can reach `github.com`. If you are behind an
> institutional firewall or VPN, configure your proxy first (see
> *Troubleshooting* below) or use the **local source** method instead.

```{r install-github, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Check GitHub is reachable before attempting install
.github_reachable <- tryCatch({
    con <- url("https://api.github.com", open = "r")
    close(con)
    TRUE
}, warning = function(w) FALSE,
   error   = function(e) FALSE)

if (!.github_reachable) {
    stop(
        "Cannot reach github.com.\n",
        "Options:\n",
        "  1. Check your internet / VPN / firewall.\n",
        "  2. Set a proxy: Sys.setenv(https_proxy = 'http://proxy:port')\n",
        "  3. Download the ZIP from https://github.com/safarafique/GExPipe\n",
        "     and install locally:\n",
        "     install.packages('path/to/GExPipe-main.zip', repos = NULL, type = 'source')"
    )
}

# dependencies = TRUE installs BOTH Imports: and Suggests: packages
# (Boruta, cicerone, mixOmics, xgboost, rms, rmda, biomaRt, etc.).
# Without this flag those optional packages are skipped and several app
# features (ML tab, nomogram, guided tour) load but silently do nothing.
# BiocManager handles CRAN + Bioconductor repos automatically — no need
# for a separate remotes call.
#
# --no-staged-install prevents a Windows file-locking bug where R's staged
# installation mechanism leaves a corrupt GExPipe.rdb file, causing
# "lazy-load database is corrupt" on the very next library(GExPipe) call.
# force = TRUE ensures the latest GitHub code installs even when an older
# version of GExPipe is already present in your library.
BiocManager::install(
    "safarafique/GExPipe",
    dependencies = TRUE,
    INSTALL_opts = "--no-staged-install",
    force        = TRUE
)

# After install, launch the app with one line.
# The R console shows a live package loading list before the app opens —
# see "Launch the Shiny app" below for all options including Google Colab.
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

### Troubleshooting: GitHub not reachable

| Situation | Fix |
|----|----|
| Corporate firewall / proxy | `Sys.setenv(https_proxy = "http://yourproxy:port")` then retry |
| VPN required | Connect to VPN first, then run the install block |
| No internet at all | Download ZIP from GitHub on another machine, copy the file, then: `install.packages("GExPipe-main.zip", repos = NULL, type = "source")` |
| `BiocManager::install` fails | Try `remotes::install_github("safarafique/GExPipe", dependencies = TRUE, INSTALL_opts = "--no-staged-install")` as a fallback |

### Troubleshooting: Windows — packages "in use and will not be installed"

This happens when packages are already loaded in your R session (Windows
locks their DLL files). Run the install in a fresh subprocess instead:

```{r fix-dll-lock, eval=FALSE}
# Runs in a clean Rscript process — nothing is loaded, no DLL locks.
system2(
  file.path(R.home("bin"), "Rscript"),
  args = c("--vanilla", "--no-save", "-e", paste0(
    'options(repos = c(CRAN = "https://cloud.r-project.org"));',
    'if (!requireNamespace("BiocManager", quietly=TRUE))',
    '  install.packages("BiocManager");',
    'BiocManager::install(',
    '  c("ggplot2","xgboost","biomaRt","Matrix","rlang","cli"),',
    '  force=TRUE, ask=FALSE, update=FALSE)'
  )),
  stdout = "", stderr = ""
)
# After it finishes, restart R (Ctrl+Shift+F10 in RStudio) then launch:
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

# Launch the Shiny app

> **R version requirement:** Use a **stable R release** (R 4.5.x or
> 4.6.x stable). Pre-release builds such as R 4.6.0 alpha have no
> pre-built package binaries — many packages fail to compile or load.
> Download the latest stable R from
> <https://cran.r-project.org/bin/windows/base/>.

GExPipe can be launched in three ways. Choose the one that best fits
your situation:

| Method | When to use |
|---|---|
| **Option 1 — GitHub (no install)** | First time, classroom, quick test — nothing to install |
| **Option 2 — Installed package** | Regular use — starts in seconds every time |
| **Option 3 — Google Colab** | No local R, shared machine, cloud analysis |

------------------------------------------------------------------------

## Option 1 — Run directly from GitHub (simplest, no install needed)

Paste **these 2 lines** into R or RStudio on any machine. Everything —
BiocManager, all Bioconductor + CRAN packages — installs and loads
automatically.

```{r launch-github, eval=FALSE}
if (!requireNamespace("shiny", quietly = TRUE)) install.packages("shiny")
# ref = "main"           → always pulls the latest committed code from the main branch
# subdir = "inst/shinyapp" → points shiny to where app.R actually lives in the repo
shiny::runGitHub(
  "GExPipe", "safarafique",
  ref    = "main",
  subdir = "inst/shinyapp",
  destdir = tempfile()
)
```

> **First run on a fresh machine** takes **10–30 minutes** (downloading
> \~50 packages). Every run after that starts in **seconds** because the
> packages are already cached in your R library.

What happens under the hood:

1.  `shiny::runGitHub()` downloads the latest commit on `main` from
    `github.com/safarafique/GExPipe` into a temporary folder and enters
    the `inst/shinyapp/` subdirectory (where `app.R` lives).
2.  `global.R` runs automatically and installs every missing dependency
    via a background R subprocess (avoids Windows DLL locks).
3.  The R console shows a live per-package loading list — `✓ loaded` or
    `✗ MISSING` for each of the 54 packages.
4.  Once all packages are ready the app opens directly to
    **Step 1 — Download Data**.

------------------------------------------------------------------------

## Option 2 — Installed package (recommended for regular use)

**First time only** — installs GExPipe and all dependencies:

```{r install-then-launch, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# dependencies = TRUE installs Imports: AND Suggests: (Boruta, cicerone,
# mixOmics, xgboost, rms, rmda, biomaRt, etc.) so every app feature works.
# force = TRUE ensures the latest GitHub code installs even when an older
# version of GExPipe is already present in your library.
BiocManager::install(
    "safarafique/GExPipe",
    dependencies = TRUE,
    INSTALL_opts = "--no-staged-install",
    force        = TRUE
)
```

**Every session after that — just one line:**

```{r launch-app, eval=FALSE}
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

What happens when you run it:

1.  `runGExPipe()` scans all 54 dependencies — takes under one second
    when everything is installed.
2.  Any missing package is installed in a **background Rscript process**
    (avoids Windows DLL locks — no session restart needed).
3.  The R console prints a live loading list before the browser opens:
    ```
    [ 1/54] shiny              ... ✓ 1.8.1
    [ 2/54] shinydashboard     ... ✓ 0.7.2
    ...
    [54/54] xgboost            ... ✓ 2.1.4
    ✓ Status: All packages ready — opening app...
    ```
4.  `shiny::runApp()` starts the server and opens the app directly to
    **Step 1 — Download Data**. The R session stays blocked until you
    stop the app (RStudio Stop button or Ctrl+C).

> **Note:** `runGExPipe()` follows the Bioconductor convention — it
> returns a Shiny app object; `shiny::runApp()` starts the actual server.

------------------------------------------------------------------------

## Option 3 — Google Colab (cloud / no local R required)

Colab has no local browser, so the app must listen on all interfaces
(`host = "0.0.0.0"`) and be exposed through a tunnel.

### Step A — Switch Colab to the R runtime (recommended)

In Colab: **Runtime → Change runtime type → R**, then run these cells:

**Cell 1 — Start an ngrok tunnel** (get a free token at
<https://ngrok.com>, takes 30 seconds to sign up):

```{r colab-r-cell1, eval=FALSE}
NGROK_TOKEN <- "PASTE_YOUR_TOKEN_HERE"   # free at ngrok.com

system(paste0(
  "wget -q https://bin.equinox.io/c/bNyj1mQVY4c/",
  "ngrok-v3-stable-linux-amd64.tgz -O ngrok.tgz",
  " && tar xzf ngrok.tgz"
), wait = TRUE)
system(paste0("./ngrok config add-authtoken ", NGROK_TOKEN), wait = TRUE)
system("./ngrok http 3838 > /dev/null 2>&1 &", wait = FALSE)
Sys.sleep(3)

url <- system(paste0(
  "curl -s http://localhost:4040/api/tunnels | ",
  "python3 -c \"import sys,json; t=json.load(sys.stdin)['tunnels']; ",
  "print(t[0]['public_url'] if t else 'ngrok not ready')\""
), intern = TRUE)
cat("\n===================================================\n")
cat("  GExPipe URL:", url, "\n")
cat("  Open after the next cell finishes installing.\n")
cat("===================================================\n")
```

**Cell 2 — Run GExPipe** (this cell blocks while the app is alive —
that is normal):

```{r colab-r-cell2, eval=FALSE}
options(timeout = 3600)   # allow long downloads

if (!requireNamespace("shiny", quietly = TRUE))
  install.packages("shiny", repos = "https://cloud.r-project.org",
                   quiet = TRUE)

# First run: installs ~50 packages automatically (15–30 min).
# Every subsequent Colab session starts in seconds.
# ref + subdir ensure the latest code is pulled from the correct app directory.
shiny::runGitHub(
  "GExPipe", "safarafique",
  ref            = "main",
  subdir         = "inst/shinyapp",
  destdir        = tempfile(),
  launch.browser = FALSE,
  host           = "0.0.0.0",
  port           = 3838L
)
```

### Step B — Alternative: Python runtime

If you prefer to keep the default Python runtime, use these cells
instead:

**Cell 1 — Install R:**

```{python colab-py-cell1, eval=FALSE}
import subprocess
subprocess.run([
    "apt-get", "install", "-y",
    "r-base", "r-base-dev",
    "libcurl4-openssl-dev", "libssl-dev", "libxml2-dev",
    "libfontconfig1-dev", "libharfbuzz-dev", "libfribidi-dev",
    "libfreetype6-dev", "libpng-dev", "libtiff5-dev", "libjpeg-dev"
], capture_output=True)
print("R installed:", subprocess.check_output(
    ["R", "--version"]).decode().split("\n")[0])
```

**Cell 2 — Start ngrok tunnel:**

```{python colab-py-cell2, eval=FALSE}
NGROK_TOKEN = "PASTE_YOUR_TOKEN_HERE"   # free at ngrok.com

import subprocess
subprocess.run(["pip", "install", "pyngrok", "-q"])
from pyngrok import ngrok
ngrok.set_auth_token(NGROK_TOKEN)
tunnel = ngrok.connect(3838)
print("=" * 55)
print(f"  GExPipe URL: {tunnel.public_url}")
print("  Open after the next cell finishes installing.")
print("=" * 55)
```

**Cell 3 — Run GExPipe** (blocks while the app is alive):

```{python colab-py-cell3, eval=FALSE}
import subprocess
subprocess.run(["Rscript", "-e", """
options(timeout = 3600)
if (!requireNamespace("shiny", quietly = TRUE))
  install.packages("shiny", repos = "https://cloud.r-project.org",
                   quiet = TRUE)
shiny::runGitHub(
  "GExPipe", "safarafique",
  ref            = "main",
  subdir         = "inst/shinyapp",
  destdir        = tempfile(),
  launch.browser = FALSE,
  host           = "0.0.0.0",
  port           = 3838L
)
"""])
```

### What to expect in Colab

| Phase | What you see | Typical time |
|---|---|---|
| Cell running, install output scrolling | Packages downloading | 15–30 min (first run) |
| `Listening on http://0.0.0.0:3838` | App is ready | — |
| Open the ngrok URL | GExPipe dashboard | — |
| Subsequent Colab sessions (same VM) | Packages already cached | \~30 sec |

> **Tip:** Colab recycles the VM after \~12 hours or on disconnect.
> Packages must be reinstalled in a new VM — GExPipe's auto-installer
> handles this automatically; just re-run the cells.

------------------------------------------------------------------------

## Advanced: pre-install or update packages separately

`gexpipe_setup()` installs or updates all dependencies without launching
the app — useful after an R upgrade or to force a full package refresh:

```{r setup-advanced, eval=FALSE}
library(GExPipe)
gexpipe_setup(update = TRUE, optional = TRUE, launch = FALSE)

# Then launch normally
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

# Workflow Example (evaluated)

This section demonstrates three core tasks with bundled data:

-   loading expression and metadata tables,
-   running basic processing suitable for downstream analysis,
-   visualizing results (PCA and expression heatmap).

## Load package and example data

```{r load-data}
library(GExPipe)
library(ggplot2)

expr_path <- system.file("extdata", "vignette_expression.csv", package = "GExPipe")
meta_path <- system.file("extdata", "vignette_sample_metadata.csv", package = "GExPipe")

stopifnot(nzchar(expr_path), file.exists(expr_path))
stopifnot(nzchar(meta_path), file.exists(meta_path))

expr_tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta_tab <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(expr_tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- expr_tab[[1]]

rownames(meta_tab) <- meta_tab$SampleID
meta_tab <- meta_tab[colnames(expr), , drop = FALSE]

dim(expr)
head(meta_tab)
```

## Process data with core pipeline helpers

`gexp_normalize_and_intersect()` accepts a named list of per-dataset
expression matrices (`micro_expr_list` for microarray / pre-normalized
data, `rna_counts_list` for raw RNA-seq counts) and returns a list with:

-   **`combined_expr`** — a single genes × samples matrix after
    within-dataset normalization and global quantile normalization
    across all datasets.
-   **`common_genes`** — the character vector of genes present in every
    dataset.
-   **`combined_expr_before_global`** — the combined matrix before the
    final global quantile step (useful for QC comparisons).

```{r process-data}
if (!"Dataset" %in% colnames(meta_tab)) {
  meta_tab$Dataset <- "D1"
}

# Build a named list: one matrix per dataset (genes × samples)
datasets <- unique(meta_tab$Dataset)
micro_expr_list <- setNames(
  lapply(datasets, function(ds) {
    sample_ids <- rownames(meta_tab)[meta_tab$Dataset == ds]
    expr[, sample_ids, drop = FALSE]
  }),
  datasets
)

norm_res <- gexp_normalize_and_intersect(
  micro_expr_list    = micro_expr_list,
  rna_counts_list    = list(),
  micro_norm_method  = "quantile",
  rnaseq_norm_method = "TMM",
  de_method          = "limma"
)

cat("Combined matrix dimensions:", dim(norm_res$combined_expr), "\n")
cat("Common genes across datasets:", length(norm_res$common_genes), "\n")
```

## Visualize processed data

### PCA plot

```{r pca-plot}
# Remove near-zero-variance genes before scaling to avoid NaN in prcomp
expr_for_pca <- norm_res$combined_expr
keep_var <- apply(expr_for_pca, 1, var, na.rm = TRUE) > 1e-8
pca <- prcomp(t(expr_for_pca[keep_var, ]), center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample  = rownames(pca$x),
  PC1     = pca$x[, 1],
  PC2     = pca$x[, 2],
  Dataset = meta_tab[rownames(pca$x), "Dataset", drop = TRUE],
  stringsAsFactors = FALSE
)

var_pct <- round(summary(pca)$importance[2, 1:2] * 100, 1)

ggplot(pca_df, aes(PC1, PC2, color = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title    = "PCA of bundled example data",
    subtitle = "After normalization and gene intersection",
    x        = paste0("PC1 (", var_pct[1], "% variance)"),
    y        = paste0("PC2 (", var_pct[2], "% variance)")
  )
```

### Expression heatmap

```{r heatmap-plot}
top_var <- order(apply(norm_res$combined_expr, 1, var, na.rm = TRUE),
                 decreasing = TRUE)
top_var <- head(top_var, min(30L, length(top_var)))

if (length(top_var) > 1L) {
  pheatmap::pheatmap(
    norm_res$combined_expr[top_var, , drop = FALSE],
    scale         = "row",
    show_rownames = length(top_var) <= 30L,
    main          = "Top variable genes (bundled example)"
  )
}
```

# App walkthrough checklist

The Shiny app is organized as a fixed 15-step pipeline. Each step must
be completed before the next one becomes active. The sections below
describe what each step does, what inputs to provide, what outputs to
expect, and what to check before moving on.

## Step 1 — Download Data

**Purpose:** Retrieve one or more GEO datasets and build a combined
expression matrix.

**Inputs:**

-   *Platform* — choose **Microarray**, **RNA-seq**, or **Merged**
    (combine microarray and RNA-seq datasets together).
-   *GEO Accession IDs* — enter one or more GSE IDs separated by commas
    or newlines (e.g., `GSE89076, GSE50760`).

**What the app does:**

1.  Downloads the series matrix files via `GEOquery`.
2.  Maps probe IDs or Ensembl IDs to HGNC gene symbols using the
    platform annotation table.
3.  Averages probes that map to the same gene.
4.  Merges all datasets into a single genes × samples matrix keeping
    only genes present in every dataset (intersection).

**Outputs:** A status log in the sidebar confirming download success,
sample counts per dataset, and the number of common genes retained.

**Example:** Enter `GSE89076` (microarray, lung disease) and click
*Start Processing*. The log should report \~500–20 000 samples × genes
depending on the dataset.

**Tip:** If a GSE contains both GPL platforms (e.g., GPL96 and GPL570),
the app processes each separately and intersects them. Choose *Merged*
only when you intentionally want to combine microarray with RNA-seq
data.

**Before continuing:** Confirm the log shows "Download complete" and the
sample count matches the GEO record.

------------------------------------------------------------------------

## Step 2 — Quality Control

**Purpose:** Identify and remove low-quality or outlier samples before
normalization.

**Inputs:** No additional inputs are required; the app uses the matrix
from Step 1.

**What the app does:**

-   Draws **boxplots** of per-sample expression distributions (raw
    values).
-   Draws **density plots** to reveal bimodal or shifted distributions.
-   Computes **Mahalanobis distance** across the first few principal
    components to flag statistical outliers. Samples beyond a threshold
    (default *p* \< 0.001) are highlighted in red.

**Outputs:** Interactive plots and a table listing flagged outlier
samples with their Mahalanobis distance and *p*-value.

**Example:** In a 40-sample dataset you might find 2–3 samples flagged.
Click each flag to inspect its boxplot. If the sample is clearly shifted
relative to all others, tick its checkbox and click *Remove Selected*.

**Tip:** Remove genuine outliers but do not over-prune. Aim to keep all
samples whose distributions overlap with the majority. Batch effects (a
cluster shifted together) should be handled in Step 5, not here.

**Before continuing:** At least 3 samples per group must remain after
removal, or downstream statistical steps will fail.

------------------------------------------------------------------------

## Step 3 — Normalize

**Purpose:** Make expression values comparable across samples and across
datasets.

**Inputs:**

-   *Mode* — **Auto** (recommended) or **Manual**.
    -   Auto applies **quantile normalization** to microarray data and
        **TMM** (edgeR) to RNA-seq counts.
    -   Manual lets you choose from quantile, log2, VST, RLE, or upper
        quantile.
-   *Log-transform* checkbox — pre-checked for microarray; leave
    unchecked for raw RNA-seq counts (TMM handles this internally).

**What the app does:** Applies within-dataset normalization, then a
second global quantile normalization across all datasets when more than
one GSE is present. This two-step approach removes platform-level
offsets while preserving within-dataset biological variation.

**Outputs:** Before/after boxplots and density overlays confirming that
all sample distributions align after normalization.

**Example:** After quantile normalization of two microarray datasets
(GSE89076 + GSE63514), all 40 boxes should align at the same median.

**Before continuing:** Verify the after-normalization boxplot shows
medians at roughly the same level across all samples.

------------------------------------------------------------------------

## Step 4 — Assign Groups

**Purpose:** Label each sample as *Normal*, *Disease*, or *Exclude* to
define the comparison for differential expression.

**Inputs:**

-   *Phenotype column* — choose the metadata column that contains
    condition labels (e.g., `characteristics_ch1`, `title`, or
    `source_name_ch1`). The dropdown is populated from the GEO series
    matrix automatically.
-   *Per-sample assignment table* — the app shows each unique metadata
    value; map it to Normal, Disease, or Exclude using dropdown
    selectors.

**What the app does:** Stores a `group` factor (Normal / Disease) for
every sample that is passed to the DE models in Step 6.

**Example:** For a GSE with values `"healthy control"` and
`"COPD patient"`, map `"healthy control"` → Normal and `"COPD patient"`
→ Disease. Mark any QC-failed samples as Exclude.

**Tip:** The Exclude option is useful for samples that passed QC
numerically but are known to be mis-labeled in GEO metadata.

**Before continuing:** The summary bar should show ≥ 3 Normal and ≥ 3
Disease samples.

------------------------------------------------------------------------

## Step 5 — Batch Correction

**Purpose:** Remove technical variation introduced by different
experiments, labs, or sequencing runs.

**Inputs:**

-   *Method* — **ComBat** (parametric empirical Bayes, robust for small
    datasets) or **limma removeBatchEffect** (linear model, slightly
    faster).
-   *Batch variable* — automatically set to the GSE ID column when
    multiple datasets are present.

**What the app does:** Applies the chosen method to regress out batch
effects while preserving the Normal/Disease contrast. Produces PCA plots
before and after correction.

**Note:** When only one GSE is loaded, Step 5 is automatically skipped
and the app proceeds directly to Step 6.

**Example:** After merging GSE89076 and GSE63514, the pre-correction PCA
shows samples clustering by dataset rather than condition. After ComBat,
samples should cluster by Normal/Disease instead.

**Tip:** If the post-correction PCA still shows dataset separation,
increase the number of PCs displayed or switch methods. A perfect
separation between batches that fully overlaps the biological signal
cannot be corrected without confounding the DE results.

**Before continuing:** The post-correction PCA should show Normal and
Disease samples intermingled across datasets.

------------------------------------------------------------------------

## Step 6 — Differential Expression

**Purpose:** Identify genes that are significantly up- or down-regulated
between Normal and Disease groups.

**Inputs:**

-   *Method* — **limma** (microarray / pre-normalized), **limma-voom**
    (RNA-seq counts), **DESeq2** (RNA-seq counts), or **edgeR** (RNA-seq
    counts).
-   *Adjusted p-value cutoff* — default 0.05 (Benjamini–Hochberg FDR).
-   *\|log₂FC\| cutoff* — default 1.0 (twofold change).

**What the app does:** Fits a linear model (limma/voom) or
negative-binomial model (DESeq2/edgeR) for the Normal vs. Disease
contrast. Applies FDR correction. Returns a ranked results table.

**Outputs:**

-   **Volcano plot** — x-axis log₂FC, y-axis −log₁₀(adj. *p*). Genes
    above the horizontal line and outside the vertical lines are
    significant DEGs.
-   **Heatmap** — top significant DEGs clustered by sample and gene.
-   **Results table** — downloadable CSV with gene, log₂FC, *p*-value,
    adj. *p*-value.

**Example:** With GSE89076 (COPD vs. controls, limma), expect \~200–800
DEGs at FDR \< 0.05 and \|log₂FC\| \> 1. Genes such as *MUC5B*, *SFTPC*,
or *AGER* commonly appear in lung-disease datasets.

**Tip:** If fewer than 20 DEGs are returned, relax the FC cutoff to 0.5
or the FDR to 0.1. If tens of thousands of genes are significant, your
dataset may have a very strong signal or a normalization issue — review
Step 3.

**Before continuing:** Download the results table and check that log₂FC
signs are biologically plausible (disease-up genes should have positive
FC if Disease is the numerator).

------------------------------------------------------------------------

## Step 7 — WGCNA (Co-expression Network)

**Purpose:** Group genes into co-expression modules and correlate each
module with the Normal/Disease trait to find biologically coherent gene
sets.

**Inputs:**

-   *Soft threshold (power β)* — WGCNA plots R² vs. β; choose the
    smallest β where scale-free topology R² ≥ 0.80 (typically β = 6–12
    for microarray, 4–8 for RNA-seq). The app highlights the recommended
    value.
-   *Minimum module size* — default 30 genes. Increase for cleaner
    modules; decrease if no modules are found.
-   *Merge cut height* — default 0.25. Lower values merge more modules.

**What the app does:** Constructs a weighted correlation network,
performs hierarchical clustering, uses `dynamicTreeCut` to detect
modules (each colored), and correlates module eigengenes with the trait
vector.

**Outputs:**

-   **Scale-free topology plot** — use this to pick β.
-   **Dendrogram + module color bar** — shows which genes belong to
    which module.
-   **Module–trait heatmap** — Pearson r and *p*-value for each module
    vs. Normal/Disease. Modules with \|r\| \> 0.5 and *p* \< 0.05 are
    considered trait-associated.

**Example:** In a COPD dataset you might find a "turquoise" module (r =
0.72, *p* = 0.001) positively correlated with disease and a "blue"
module (r = −0.65) negatively correlated.

**Tip:** WGCNA requires ≥ 15 samples. With fewer samples, reduce minimum
module size to 10 and accept that modules may be less reliable.

**Before continuing:** At least one module should have a significant
trait correlation (*p* \< 0.05).

------------------------------------------------------------------------

## Step 8 — Common Genes (DEG ∩ WGCNA Hub Genes)

**Purpose:** Intersect DEGs from Step 6 with hub genes from the
trait-associated WGCNA modules to produce a high-confidence candidate
gene list.

**Inputs:**

-   *Module selection* — tick one or more trait-correlated modules from
    the WGCNA output.
-   *Hub gene criterion* — genes in the top X% of intramodular
    connectivity (default top 10%) are considered hub genes.

**What the app does:** Computes the intersection of the DEG list with
the selected module hub genes. Runs Gene Ontology (GO BP, MF, CC) and
KEGG pathway over-representation analysis (ORA) on the intersection
using `clusterProfiler`.

**Outputs:**

-   Venn/UpSet diagram showing DEG count, hub gene count, and
    intersection.
-   GO and KEGG dot plots ranked by adjusted *p*-value.
-   Downloadable table of common genes with their log₂FC, module
    membership, and connectivity score.

**Example:** 500 DEGs ∩ 150 turquoise hub genes → 42 common genes
enriched in "cilium assembly" (<GO:0060271>) and "MAPK signaling
pathway" (hsa04010).

**Tip:** If the intersection is empty, relax the hub gene criterion
(increase from top 10% to top 25%) or select an additional module.

**Before continuing:** Keep a note of the common gene count — this list
feeds Steps 9–15.

------------------------------------------------------------------------

## Step 9 — PPI Network

**Purpose:** Map the common genes onto a protein–protein interaction
network to identify the most connected (hub) proteins.

**Inputs:**

-   *STRINGdb confidence score cutoff* — default 400 (medium
    confidence). Increase to 700 for high-confidence interactions only.
-   *Network layout* — choose Fruchterman–Reingold, Kamada–Kawai, or
    circle.

**What the app does:** Queries `STRINGdb` for interactions among the
common genes. Builds an `igraph` network. Ranks nodes by degree (number
of interactions), betweenness centrality, and eigenvector centrality.
The top 10 hub proteins are highlighted.

**Outputs:**

-   Interactive network plot — node size proportional to degree; color
    indicates centrality.
-   Hub gene table with degree, betweenness, and eigenvector scores
    (downloadable CSV).

**Example:** From 42 common genes in a COPD network, hub proteins such
as *TP53*, *AKT1*, *EGFR*, or *TNF* typically emerge as highly connected
nodes.

**Tip:** A minimum of \~10 nodes with at least one edge is required to
render a meaningful network. If the network is empty, lower the STRING
confidence cutoff to 200.

**Before continuing:** Note the top 5 hub genes by degree — these are
the primary candidates for Steps 10–14.

------------------------------------------------------------------------

## Step 10 — Machine-Learning Feature Selection

**Purpose:** Use multiple ML algorithms to identify the smallest set of
genes that best distinguishes Normal from Disease samples.

**Inputs:**

-   *Algorithms* — select one or more: **LASSO**, **Elastic Net**,
    **Ridge**, **Random Forest**, **SVM-RFE**, **Boruta**, **sPLS-DA**,
    **XGBoost + SHAP**.
-   *Cross-validation folds* — default 10-fold CV.
-   *SHAP values* — automatically computed for XGBoost; shows which
    genes drive each prediction.

**What the app does:** Trains each selected model on the normalized
expression of the common genes. Each algorithm returns a ranked
importance list or a selected-feature subset. The final feature list is
the intersection (or union) of genes selected by ≥ 2 algorithms.

**Outputs:**

-   Per-algorithm importance bar charts.
-   SHAP summary plot (if XGBoost selected).
-   Final candidate gene table — genes consistently selected across
    algorithms.

**Example:** Running LASSO + Random Forest + SVM-RFE on 42 common genes
might converge on 6–10 genes as the most discriminative features.

**Tip:** Start with LASSO and Random Forest for speed. Add SVM-RFE and
XGBoost for robustness if computation time allows. On a standard laptop,
42 genes × 10-fold CV finishes in under 1 minute per algorithm.

**Before continuing:** Select 3–15 final candidate genes to carry
forward. Too many genes reduce nomogram interpretability.

------------------------------------------------------------------------

## Step 11 — Validation

**Purpose:** Confirm that the candidate genes generalize beyond the
training data.

**Inputs:**

-   *Validation type*:
    -   **External** — provide a new independent GEO GSE ID. The app
        downloads and normalizes it automatically, then scores samples
        using the training-set model.
    -   **Internal** — randomly splits the current dataset 70 % training
        / 30 % test (stratified by group).
-   *External GSE* — if External is chosen, enter the new GSE ID (e.g.,
    `GSE47460`).

**What the app does:** Applies the model fitted in Step 10 to the
validation set. Reports accuracy, sensitivity, specificity, and
confusion matrix.

**Example:** Training on GSE89076 (COPD) and validating on GSE47460
(independent COPD cohort) gives external AUC that should exceed 0.75 for
the results to be publishable.

**Tip:** Always prefer external validation when an independent cohort
exists in GEO for the same disease. Internal validation is acceptable as
a secondary confirmation.

**Before continuing:** Validation accuracy ≥ 70 % is a reasonable
minimum before proceeding to ROC analysis.

------------------------------------------------------------------------

## Step 12 — ROC Analysis

**Purpose:** Quantify the diagnostic accuracy of each candidate gene
individually using Receiver Operating Characteristic curves.

**Inputs:**

-   *AUC filter* — default ≥ 0.80. Genes below this threshold are
    excluded from the nomogram and GSEA steps.
-   *Training vs. Validation* — toggle between datasets.

**What the app does:** For each candidate gene, fits a simple logistic
regression, computes the ROC curve using `pROC`, and reports AUC with 95
% confidence interval. Overlays all curves on one plot.

**Outputs:**

-   Multi-curve ROC plot colored by gene.
-   AUC table with CI (downloadable).

**Example:** From 8 candidate genes, suppose 5 have AUC ≥ 0.80 in both
training and validation. These 5 genes proceed to the nomogram.

**Tip:** A gene with AUC ≥ 0.80 in training but \< 0.65 in validation is
likely overfitted — remove it.

**Before continuing:** Retain only genes with AUC ≥ 0.80 in both cohorts
(or relax to ≥ 0.70 if the cohort is small).

------------------------------------------------------------------------

## Step 13 — Nomogram

**Purpose:** Build a clinically interpretable scoring model that
predicts disease probability from the final gene panel.

**Inputs:**

-   *Genes included* — the genes that passed ROC filtering (auto-filled
    from Step 12).
-   *Calibration bootstrap replications* — default 200.

**What the app does:**

1.  Fits a binary logistic regression with `rms::lrm()`.
2.  Draws a **nomogram** — each gene has a scale; a clinician sums
    points across genes to read off a predicted probability.
3.  Draws a **calibration curve** — predicted vs. observed probability;
    a curve close to the diagonal is well-calibrated.
4.  Runs **Decision Curve Analysis (DCA)** via `rmda` — shows net
    benefit across decision thresholds, comparing the model against
    treat-all and treat-none strategies.

**Outputs:** Nomogram plot, calibration curve, and DCA plot (all
downloadable as PNG or PDF).

**Example:** A 4-gene nomogram for COPD might have total points mapping
to disease probabilities of 5–95 %. A calibration curve with mean
absolute error \< 0.05 indicates excellent calibration.

**Tip:** If calibration is poor (curve deviates \> 0.10 from diagonal),
reduce the number of genes in the model or increase bootstrap
replications.

**Before continuing:** DCA net benefit should be positive over a
clinically relevant threshold range (typically 10–40 % for disease
screening).

------------------------------------------------------------------------

## Step 14 — GSEA

**Purpose:** Determine which biological pathways are enriched in samples
with high vs. low expression of each final candidate gene.

**Inputs:**

-   *Gene set collection* — **GO** (BP, MF, CC), **KEGG**, or **MsigDB**
    (Hallmark, C2 curated, C5 ontology, etc.).
-   *p-value cutoff* — default 0.05 (BH-adjusted).
-   *Target gene* — run GSEA ranked by correlation with one gene at a
    time, or use all genes in the panel together.

**What the app does:** Ranks all genes in the expression matrix by their
Pearson correlation with the target gene's expression (or by signed
log₂FC for the DE ranking). Runs `clusterProfiler::gseGO()` or
`gseKEGG()` on the ranked list.

**Outputs:**

-   **Ridge plot** — enrichment score distribution per pathway.
-   **GSEA enrichment plot** — running enrichment score for the top
    pathway.
-   Results table with NES, *p*-value, and leading-edge genes
    (downloadable).

**Example:** For a hub gene such as *SFTPC* in a COPD dataset, GSEA
(KEGG) might show "ECM-receptor interaction" (NES = −1.8, *p* = 0.002)
as the top negatively enriched pathway.

**Tip:** Use MsigDB Hallmark for a high-level mechanistic overview; use
KEGG for pathway-specific drug target identification.

**Before continuing:** Note the top 3 enriched pathways per gene — they
contextualize the biology section of the final report.

------------------------------------------------------------------------

## Step 15 — Summary Report

**Purpose:** Generate a structured narrative and a downloadable PDF
summarizing the entire analysis.

**Inputs:** No additional inputs. The report is assembled from results
stored in all previous steps.

**What the app does:** Writes an auto-generated paragraph covering
dataset details (GEO IDs, sample counts), number of DEGs, WGCNA module
findings, common gene count, PPI hub genes, ML-selected features,
validation AUCs, and the top GSEA pathways. Renders the paragraph to PDF
using `rmarkdown`.

**Outputs:**

-   Narrative text displayed in the app panel.
-   *Download PDF* button — produces a formatted report suitable for
    inclusion in a manuscript Methods section.

**Example output paragraph:**

> "Two microarray datasets (GSE89076, n = 40; GSE63514, n = 48) were
> retrieved from GEO. After QC, normalization (quantile), and ComBat
> batch correction, 312 DEGs (FDR \< 0.05, \|log₂FC\| \> 1) were
> identified by limma. WGCNA detected 8 modules; the turquoise module (r
> = 0.72, *p* \< 0.001) was most strongly associated with COPD. The
> intersection of DEGs and turquoise hub genes yielded 42 common genes.
> After PPI analysis and multi-algorithm ML feature selection (LASSO,
> RF, SVM-RFE), 6 final biomarkers were identified: *SFTPC*, *MUC5B*,
> *AGER*, *CLDN18*, *LAMP3*, *NAPSA* (validation AUC range 0.81–0.94).
> GSEA revealed significant enrichment of ECM-receptor interaction and
> cilium assembly pathways."

**Tip:** Run Step 15 last, after all previous steps are complete.
Re-running any earlier step does not automatically regenerate the report
— click *Generate Report* again after any changes.

# App screenshots

If screenshots are placed under `inst/extdata/vignette-screenshots/`,
they are rendered below.

```{r screenshots, echo=FALSE, results='asis'}
shot_dir <- system.file("extdata", "vignette-screenshots", package = "GExPipe")

shot_meta <- list(
  list(file = "step1_download.png",  label = "Step 1 — Download Data"),
  list(file = "step2_qc.png",        label = "Step 2 — Quality Control"),
  list(file = "step6_de.png",        label = "Step 6 — Differential Expression"),
  list(file = "step9_ppi.png",       label = "Step 9 — PPI Network"),
  list(file = "step15_summary.png",  label = "Step 15 — Results Summary")
)

if (nzchar(shot_dir)) {
  for (item in shot_meta) {
    p <- file.path(shot_dir, item$file)
    if (file.exists(p)) {
      cat("\n\n## ", item$label, "\n\n", sep = "")
      print(knitr::include_graphics(p))
    }
  }
}
```

# Session Information

```{r sessionInfo}
sessionInfo()
```

1
