---
title: "GExPipe Shiny app in R/ and automated testing"
author: "Safa Rafique, Naeem Mehmood Ashraf, Farooq Sabir"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
package: GExPipe
vignette: >
  %\VignetteIndexEntry{GExPipe Shiny app in R/ and automated testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# Overview

Bioconductor [expects Shiny software packages](https://contributions.bioconductor.org/shiny.html) to keep **UI and server logic under `R/`**, to **return** a `shiny.appobj` from an exported function (without calling `shiny::runApp()` inside package code), and to provide **automated tests**—including **`shinytest2`** for app smoke or deeper checks where feasible.

`GExPipe` follows that layout: the exported entrypoint is `runGExPipe()`, which returns an app built from internal `gexp_app_ui()` and `gexp_app_server()` defined in `R/`. Step-specific UI and server code is loaded from `inst/shinyapp/` (and `R/shiny_src/`) via thin wrappers so the **orchestration** lives in package R code, consistent with the recommended `interface_*`, `server_*`, `observers_*`, and `utils_*` naming style where applicable.

# Runnable: build the app object (no browser)

The chunk below runs during `R CMD build` / vignette checks. Test-mode options limit work on the first server tick so checks stay practical; omit them for a full cold start in interactive use.

```{r build-app}
library(BiocStyle)
library(GExPipe)
library(shiny)

options(
  shiny.testmode = TRUE,
  gexpipe.minimal_attach_in_testmode = TRUE,
  gexpipe.wgcna_threads = 0L
)

app <- runGExPipe(launch.browser = FALSE, port = 0L)
stopifnot(inherits(app, "shiny.appobj"))
# Shiny >= 1.13 represents apps as httpHandler + serverFuncSource (see ?shiny::shinyApp).
stopifnot(is.function(app$httpHandler), is.function(app$serverFuncSource))
```

# Where the R code lives

| Area | Role | Example files |
|------|------|----------------|
| App entry | Return `shinyApp()` | `R/runGExPipe.R` |
| Full-page UI / dashboard shell | `gexp_app_ui()`, analysis dashboard | `R/interface_app.R`, `R/interface_welcome.R`, `R/interface_tabs_wrappers.R` |
| Server wiring | Session, `reactiveValues`, modules | `R/server_app.R`, `R/server_modules_wrappers.R` |
| Cross-cutting observers | Pipeline, navigation, workspace, help | `R/observers_*.R` |
| Shiny helpers | Attach, paths, `onStart` | `R/utils_shiny_app.R` |
| Analysis pipelines (scriptable) | Exported functions | `R/gexp_*_pipeline.R` |

Legacy UI/server sources under `R/shiny_src/` are loaded at runtime through `.gexp_inst_file()` so the same app works from an installed package.

# Testing (testthat and shinytest2)

Non-reactive helpers are covered with **testthat** under `tests/testthat/` (e.g. `test-qc-pipeline.R`, `test-batch-de-pipeline.R`).

The Shiny app is smoke-tested with **`shinytest2`** in `tests/testthat/test-shiny-integration.R`: a temporary `app.R` calls `GExPipe::runGExPipe()` and `AppDriver` verifies the session becomes ready. That test skips when `chromote` cannot start a browser (typical on some CI images). See `vignettes/gexpipe-shiny-testing-ci.Rmd` for environment variables (`CI`, `GEXPIPE_SKIP_SHINYTEST2`, `GEXPIPE_SHINYTEST2_MS`, `GEXPIPE_SHINYTEST2_USE_PKGLOAD`).

Official guidance: [Chapter 18 — Shiny apps, Testing](https://contributions.bioconductor.org/shiny.html#testing).

# Session information

```{r session-info}
sessionInfo()
```
