---
title: "GExPipe: A Pipeline for Gene Expression Analysis"
author: "Safa Rafique"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{1. Introduction to GExPipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`GExPipe` is a Shiny-first workflow for bulk RNA-seq and microarray analysis. The package was created to provide a single, guided pipeline that connects data retrieval, quality control, normalization, differential expression, network analysis, and downstream interpretation without requiring users to manually stitch many separate tools together.

Within Bioconductor, `GExPipe` builds on established infrastructure such as `r BiocStyle::Biocpkg("GEOquery")`, `r BiocStyle::Biocpkg("limma")`, `r BiocStyle::Biocpkg("DESeq2")`, `r BiocStyle::Biocpkg("edgeR")`, `r BiocStyle::Biocpkg("WGCNA")`, and `r BiocStyle::Biocpkg("clusterProfiler")`. Compared with piecemeal scripts, the app emphasizes usability and reproducibility through a fixed step-by-step interface and integrated summaries.

This vignette is user-oriented: it explains how to install, launch, and run a minimal executable workflow using package-bundled example data.

# Installation

Install `GExPipe` with `BiocManager`:

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GExPipe")
```

# Launch the Shiny app

`runGExPipe()` returns a Shiny app object. Start it with `shiny::runApp()`.

```{r launch-app, eval=FALSE}
app <- GExPipe::runGExPipe(launch.browser = TRUE, port = 3838L)
shiny::runApp(app, port = 3838L)
```

# Workflow Example (evaluated)

This section demonstrates three core tasks with bundled data:

- loading expression and metadata tables,
- running basic processing suitable for downstream analysis,
- visualizing results (PCA and expression heatmap).

## Load package and example data

```{r load-data}
library(GExPipe)
library(ggplot2)

expr_path <- system.file("extdata", "vignette_expression.csv", package = "GExPipe")
meta_path <- system.file("extdata", "vignette_sample_metadata.csv", package = "GExPipe")

stopifnot(nzchar(expr_path), file.exists(expr_path))
stopifnot(nzchar(meta_path), file.exists(meta_path))

expr_tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta_tab <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(expr_tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- expr_tab[[1]]
colnames(expr) <- gsub("^X", "", colnames(expr))

rownames(meta_tab) <- meta_tab$SampleID
meta_tab <- meta_tab[colnames(expr), , drop = FALSE]

dim(expr)
head(meta_tab)
```

## Process data with core pipeline helpers

```{r process-data}
if (!"Dataset" %in% colnames(meta_tab)) {
  meta_tab$Dataset <- "D1"
}

micro_expr_list <- split(
  x = as.data.frame(expr, check.names = FALSE),
  f = meta_tab$Dataset
)
micro_expr_list <- lapply(micro_expr_list, function(df) as.matrix(df))

# Rebuild per-dataset matrices with genes as rows and samples as columns.
for (ds in names(micro_expr_list)) {
  sample_ids <- rownames(meta_tab)[meta_tab$Dataset == ds]
  micro_expr_list[[ds]] <- expr[, sample_ids, drop = FALSE]
}

norm_res <- gexp_normalize_and_intersect(
  micro_expr_list = micro_expr_list,
  rna_counts_list = list(),
  micro_norm_method = "quantile",
  rnaseq_norm_method = "TMM",
  de_method = "limma"
)

dim(norm_res$combined_expr)
length(norm_res$common_genes)
```

## Visualize processed data

```{r pca-plot}
pca <- prcomp(t(norm_res$combined_expr), center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample = rownames(pca$x),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  Dataset = meta_tab[rownames(pca$x), "Dataset", drop = TRUE],
  stringsAsFactors = FALSE
)

ggplot(pca_df, aes(PC1, PC2, color = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "PCA of bundled example data",
    subtitle = "After normalization and gene intersection"
  )
```

```{r heatmap-plot}
top_var <- order(apply(norm_res$combined_expr, 1, var, na.rm = TRUE), decreasing = TRUE)
top_var <- head(top_var, min(30L, length(top_var)))

if (length(top_var) > 1L) {
  pheatmap::pheatmap(
    norm_res$combined_expr[top_var, , drop = FALSE],
    scale = "row",
    main = "Top variable genes (bundled example)"
  )
}
```

# App walkthrough checklist

For real analyses in the Shiny app, use this sequence:

1. Step 1: select platform and GEO IDs, then start processing.
2. Step 2: review QC plots and outliers.
3. Step 3-5: normalize, define groups, and handle batch effects.
4. Step 6-8: run DE, WGCNA, and compute common genes.
5. Step 9-15: enrichment, PPI, ML, validation, ROC/nomogram, GSEA, and summary.

# Optional screenshots in this vignette

If screenshots are placed under `inst/extdata/vignette-screenshots/`, they are rendered below.

```{r screenshots, echo=FALSE}
shot_dir <- system.file("extdata", "vignette-screenshots", package = "GExPipe")
shot_files <- c(
  "step1_download.png",
  "step2_qc.png",
  "step6_de.png",
  "step9_ppi.png",
  "step15_summary.png"
)

if (nzchar(shot_dir)) {
  for (nm in shot_files) {
    p <- file.path(shot_dir, nm)
    if (file.exists(p)) {
      cat("\n\n## ", nm, "\n\n", sep = "")
      print(knitr::include_graphics(p))
    }
  }
}
```

# Session Information

```{r sessionInfo}
sessionInfo()
```
