---
title: "GExPipe"
author: "Safa Rafique, Naeem Mehmood Ashraf, Farooq Sabir"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: GExPipe
bibliography: ../inst/references.bib
vignette: >
  %\VignetteIndexEntry{GExPipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
set.seed(123)
```

# GExPipe

# Introduction

`GExPipe` is an **R** package and **Shiny** application for end-to-end analysis of **bulk RNA-seq** and **microarray** gene expression data, with a strong focus on studies archived in the [Gene Expression Omnibus (GEO)](https://www.ncbi.nlm.nih.gov/geo/) [@edgar2002gene]. The interactive app guides users through download, quality control, normalization, batch correction, differential expression, co-expression network analysis (WGCNA), pathway enrichment (e.g. **clusterProfiler** [@yu2012clusterprofiler]), protein–protein interaction mining, and machine learning–based biomarker exploration.

Exported **R** functions implement the same core steps so workflows can be scripted, tested, and reproduced outside the GUI—aligned with Bioconductor expectations for Shiny software packages [@chang2024shiny]. Differential expression integrates **limma** [@ritchie2015limma] and, when appropriate, **DESeq2** [@love2014moderated] and **edgeR** (via the app and pipeline helpers). WGCNA-related preparation follows standard weighted correlation network practice [@langfelder2008wgcna].

This vignette uses **bundled example matrices and sample metadata** for the full core pipeline through **WGCNA preparation**, then continues with **downstream steps** that mirror the Shiny app’s **Common genes**, **GO/KEGG**, **PPI**, **machine learning**, **ROC**, **validation**, and **GSEA** tabs (Bioconductor packages: `clusterProfiler`, `org.Hs.eg.db`, `STRINGdb`, `randomForest`, `pROC`, `msigdbr`, etc.). **STRING** and **KEGG REST** calls are wrapped in `tryCatch()` so vignette builds succeed offline; with network access you may see more enrichment and interaction hits. For GEO download examples, see the optional chunk at the end and `gexpipe-workflow.Rmd`.

# Loading packages

```{r library-import, message=FALSE}
library(BiocStyle)
library(GExPipe)
library(clusterProfiler)
library(org.Hs.eg.db)
library(randomForest)
library(pROC)
library(msigdbr)
library(STRINGdb)
```

# Bundled example data

Expression and phenotype tables live in `inst/extdata`. The helper below resolves paths both from an installed package and from a source checkout.

```{r read-extdata}
vignette_extdata_file <- function(name) {
  p <- system.file("extdata", name, package = "GExPipe")
  if (nzchar(p) && file.exists(p)) {
    return(p)
  }
  candidates <- c(
    file.path(getwd(), "..", "inst", "extdata", name),
    file.path(getwd(), "..", "..", "inst", "extdata", name),
    file.path(getwd(), "inst", "extdata", name)
  )
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0L) {
    stop("Could not find bundled vignette file: ", name)
  }
  normalizePath(hits[1], winslash = "/", mustWork = TRUE)
}

expr_path <- vignette_extdata_file("vignette_expression.csv")
meta_path <- vignette_extdata_file("vignette_sample_metadata.csv")

tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- tab[[1]]
colnames(expr) <- gsub("^X", "", colnames(expr))

rownames(meta) <- meta$SampleID
meta <- meta[colnames(expr), , drop = FALSE]

dim(expr)
head(meta)
```

# Parse GEO study identifiers

The same identifiers you would type into Step 1 of the app can be parsed programmatically (comma-separated **GSE** IDs, platform mode, DE method).

```{r parse-gse-inputs}
parsed <- gexp_parse_gse_inputs(
  analysis_type = "merged",
  rnaseq_gses = "GSE50760",
  microarray_gses = "GSE89076",
  dataset_mode = "multi"
)
parsed$rnaseq_ids
parsed$micro_ids
```

# Normalization and intersection of common genes

We split the example matrix into two synthetic “datasets”, normalize microarray-like slices, and retain the gene intersection (typical merged-platform workflow).

```{r normalize-intersect}
cols_d1 <- rownames(meta)[meta$Dataset == "D1"]
cols_d2 <- rownames(meta)[meta$Dataset == "D2"]

micro_expr_list <- list(
  D1 = expr[, cols_d1, drop = FALSE],
  D2 = expr[, cols_d2, drop = FALSE]
)
rna_counts_list <- list()

norm <- gexp_normalize_and_intersect(
  micro_expr_list = micro_expr_list,
  rna_counts_list = rna_counts_list,
  micro_norm_method = "quantile",
  rnaseq_norm_method = "TMM",
  de_method = "limma"
)

dim(norm$combined_expr)
length(norm$common_genes)
```

```{r prepare-metadata}
um <- norm$unified_metadata
um$Condition <- meta[um$SampleID, "Condition"]
um$Dataset <- meta[um$SampleID, "Dataset"]
table(um$Dataset, um$Condition)
```

# Batch correction

```{r batch-correct}
batch_res <- gexp_batch_correct(
  expr = norm$combined_expr,
  metadata = um[, c("Dataset", "Condition")],
  variance_percentile = 25,
  method = "limma"
)

dim(batch_res$expr_filtered)
dim(batch_res$batch_corrected)
```

# Differential expression

```{r run-de}
de_res <- gexp_run_de(
  expr = batch_res$batch_corrected,
  metadata = um[, c("Dataset", "Condition")],
  method = "limma",
  logfc_cutoff = 0.25,
  padj_cutoff = 0.10
)

de_df <- de_res$de_results
table(de_df$Significance)
head(de_df[, c("Gene", "logFC", "adj.P.Val", "Significance")], 8)
```

# Quality control summaries

```{r qc-helpers}
qc_out <- gexp_qc_detect_outliers(norm$combined_expr, top_n = 100)
length(qc_out$all_outliers)

qc_summary <- gexp_qc_gene_overlap_summary(
  all_genes_list = list(
    D1 = rownames(micro_expr_list$D1),
    D2 = rownames(micro_expr_list$D2)
  ),
  common_genes = norm$common_genes
)
qc_summary
```

# WGCNA preparation

```{r wgcna-prepare}
wgcna_prep <- gexp_wgcna_prepare(
  expr = batch_res$batch_corrected,
  metadata = um,
  gene_mode = "top_variable",
  top_genes = min(100L, nrow(batch_res$batch_corrected)),
  min_samples_frac = 0.5
)

dim(wgcna_prep$datExpr)
head(wgcna_prep$gene_variance_table, 3)
```

# Downstream analysis (Shiny Steps 8–15)

The table below maps **this vignette** to the **GExPipe** app tabs. All code below is **evaluated** when you knit the vignette; network-dependent steps fail gracefully when offline.

| App focus | Vignette section |
|-----------|------------------|
| **8. Common genes** (DEG ∩ WGCNA) | Common genes, overlap summary |
| **8–9. GO / KEGG** (ORA on common genes) | GO BP, MF, CC; KEGG ORA (tables) |
| **9. PPI** (STRING) | STRING interactions (optional network) |
| **10. Machine learning** | Random forest + importance table |
| **12. ROC** | AUC from forest probabilities (numeric) |
| **11. Validation** | Train on D1, predict on D2 (external-style holdout) |
| **14. GSEA** | `gseGO` / `gseKEGG` tables on ranked DE statistics |
| **MsigDB / Hallmark** | Gene set tables via `msigdbr` |

# Common genes (DEG and WGCNA overlap)

In the app (**Step 8**), **common genes** are the intersection of **significant DEGs** (differential expression) and genes that fall in **trait-related WGCNA modules**. This vignette uses the same idea in compact form: significant DEGs ∩ genes kept in `gexp_wgcna_prepare()` output (`wgcna_prep$datExpr`). We also summarize overlap with **`gexp_qc_gene_overlap_summary()`** (exported helper used in QC logic).

```{r common-genes}
deg_genes <- unique(de_df$Gene[de_df$Significance != "Not Significant"])
wgcna_genes <- rownames(wgcna_prep$datExpr)
common_genes <- intersect(deg_genes, wgcna_genes)

if (length(common_genes) < 5L) {
  deg_genes <- head(de_df$Gene[order(de_df$adj.P.Val)], 40)
  common_genes <- intersect(deg_genes, wgcna_genes)
}

length(common_genes)
head(common_genes, 10)

gexp_qc_gene_overlap_summary(
  all_genes_list = list(DEG = deg_genes, WGCNA = wgcna_genes),
  common_genes = common_genes
)
```

The app also offers **Venn** and **UpSet** views for the same lists; this vignette sticks to **tables** only.

```{r common-genes-upset-check}
up_meta <- gexp_qc_prepare_upset_data(list(DEG = deg_genes, WGCNA = wgcna_genes))
up_meta$ok ## TRUE when both lists are non-empty (data ready for UpSet in the app)
```

# Gene Ontology enrichment (BP, MF, CC)

We map **gene symbols** to **Entrez IDs** with `clusterProfiler::bitr()`, then run **`enrichGO()`** separately for **Biological Process**, **Molecular Function**, and **Cellular Component**—the same ORA pattern as the app’s GO panels on the **common-genes** path.

```{r go-map-ids}
eg <- tryCatch(
  clusterProfiler::bitr(
    common_genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db::org.Hs.eg.db
  ),
  error = function(e) data.frame(SYMBOL = character(), ENTREZID = character())
)
nrow(eg)
```

```{r go-enrichment-all}
go_params <- list(
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  pAdjustMethod = "BH",
  pvalueCutoff = 0.2,
  qvalueCutoff = 0.3,
  readable = TRUE
)

ego_list <- if (nrow(eg) == 0L) {
  list(BP = NULL, MF = NULL, CC = NULL)
} else {
  lapply(c(BP = "BP", MF = "MF", CC = "CC"), function(ont) {
    tryCatch(
      do.call(
        clusterProfiler::enrichGO,
        c(list(gene = eg$ENTREZID, ont = ont), go_params)
      ),
      error = function(e) NULL
    )
  })
}

vapply(ego_list, function(x) if (is.null(x)) 0L else nrow(as.data.frame(x)), integer(1))
```

```{r go-tables}
for (nm in names(ego_list)) {
  ed <- if (is.null(ego_list[[nm]])) data.frame() else as.data.frame(ego_list[[nm]])
  cat("## ", nm, " — n terms:", nrow(ed), "\n", sep = "")
  if (nrow(ed) > 0L) {
    show_cols <- intersect(c("Description", "pvalue", "p.adjust", "Count"), names(ed))
    print(head(ed[, show_cols, drop = FALSE], 3))
  }
}
```

# KEGG pathway enrichment (over-representation)

`enrichKEGG()` uses the **KEGG REST API** (needs **network** access for live queries). The chunk always runs; if you are offline or the service is unreachable, the result is empty.

```{r kegg-enrichment}
ekegg <- if (nrow(eg) == 0L) {
  NULL
} else {
  tryCatch(
    clusterProfiler::enrichKEGG(
      gene = eg$ENTREZID,
      organism = "hsa",
      pvalueCutoff = 0.2,
      qvalueCutoff = 0.3
    ),
    error = function(e) NULL
  )
}

kegg_df <- if (!is.null(ekegg)) as.data.frame(ekegg) else data.frame()
nrow(kegg_df)
if (nrow(kegg_df) > 0L) {
  head(kegg_df[, intersect(c("ID", "Description", "p.adjust", "Count"), names(kegg_df))], 5)
}
```

# Protein–protein interactions (STRING)

The **PPI** tab maps your gene list to **STRING** identifiers and retrieves **experimental / database** edges above a score threshold. Below we cap the query size for speed; increase `head(common_genes, …)` for a fuller network when online.

```{r ppi-string}
ppi_res <- tryCatch(
  {
    if (length(common_genes) < 3L) {
      stop("Need at least three common genes for STRING.")
    }
    string_db <- STRINGdb$new(version = "12", species = 9606, score_threshold = 400)
    mapped <- string_db$map(
      data.frame(gene = head(common_genes, 120L)),
      "gene",
      removeUnmappedRows = TRUE
    )
    if (nrow(mapped) < 2L) {
      stop("Too few STRING mappings for this gene list.")
    }
    hits <- string_db$get_interactions(mapped$STRING_id)
    list(
      n_edges = nrow(hits),
      msg = paste0("Retrieved ", nrow(hits), " STRING interaction rows (score ≥ 400).")
    )
  },
  error = function(e) {
    list(n_edges = NA_integer_, msg = paste0("PPI skipped: ", conditionMessage(e)))
  }
)

ppi_res$msg
ppi_res$n_edges
```

# Machine learning (supervised classification)

**Step 10 (ML)** in the app trains classifiers on expression features (often **common genes**). Here we fit a **random forest** on **batch-corrected** values, **Normal vs Disease**, and show a short **importance** table when the fit succeeds.

```{r ml-rf}
genes_ml <- common_genes[seq_len(min(25L, length(common_genes)))]
sid <- colnames(batch_res$batch_corrected)
X <- t(batch_res$batch_corrected[genes_ml, sid, drop = FALSE])
y <- factor(um[sid, "Condition"], levels = c("Normal", "Disease"))

X[!is.finite(X)] <- NA
keep <- stats::complete.cases(X) & !is.na(y)
X <- X[keep, , drop = FALSE]
y <- droplevels(y[keep])

if (ncol(X) > 0L) {
  vr <- apply(X, 2L, function(z) stats::var(z, na.rm = TRUE))
  okc <- is.finite(vr) & vr > 0
  X <- if (any(okc)) X[, okc, drop = FALSE] else X[, integer(), drop = FALSE]
}

rf_fit <- if (isTRUE(nrow(X) >= 3L) && ncol(X) >= 1L && length(unique(y)) >= 2L) {
  tryCatch(
    randomForest::randomForest(x = X, y = y, ntree = 150L, importance = TRUE),
    error = function(e) NULL
  )
} else {
  NULL
}
rf_fit

if (!is.null(rf_fit) && !is.null(rf_fit$importance) && nrow(rf_fit$importance) > 0L) {
  imp <- as.data.frame(rf_fit$importance)
  imp$Gene <- rownames(imp)
  imp <- imp[order(imp$MeanDecreaseGini, decreasing = TRUE), , drop = FALSE]
  head(imp[, c("Gene", "MeanDecreaseGini"), drop = FALSE], min(12L, nrow(imp)))
}
```

# ROC curve (discrimination)

**Step 12 (ROC)** uses predicted probabilities vs true labels. We use **training-set** probabilities from the forest below **only** as a **teaching example**; for unbiased performance, use cross-validation or a held-out set (next section). Only the **AUC** is printed here (no figure).

```{r roc-demo}
roc_auc_txt <- "ROC / AUC not computed."
if (!is.null(rf_fit)) {
  oob <- tryCatch(
    predict(rf_fit, type = "prob"),
    error = function(e) NULL
  )
  if (!is.null(oob) && is.matrix(oob) && ncol(oob) >= 2L && "Disease" %in% colnames(oob)) {
    roc_obj <- tryCatch(
      pROC::roc(response = y, predictor = oob[, "Disease"], quiet = TRUE),
      error = function(e) NULL
    )
    if (!is.null(roc_obj)) {
      roc_auc_txt <- sprintf(
        "AUC (training predictions, illustrative) = %.4f",
        as.numeric(pROC::auc(roc_obj))
      )
    }
  }
}
roc_auc_txt
```

# Validation (external-style holdout)

**Step 11 (Validation)** in the app supports **external cohorts**. Here we **simulate** that by treating **Dataset D1** as **training** and **Dataset D2** as **external test**—same genes (`common_genes` subset), shared **Condition** labels—so the code path is fully **runnable** on bundled data.

```{r validation-external-style}
val_auc <- NA_real_
val_txt <- "Validation not run."

genes_val <- common_genes[seq_len(min(20L, length(common_genes)))]
sid_tr <- intersect(rownames(um)[um$Dataset == "D1"], colnames(batch_res$batch_corrected))
sid_te <- intersect(rownames(um)[um$Dataset == "D2"], colnames(batch_res$batch_corrected))

if (length(genes_val) >= 5L && length(sid_tr) >= 3L && length(sid_te) >= 2L) {
  Xtr <- t(batch_res$batch_corrected[genes_val, sid_tr, drop = FALSE])
  ytr <- droplevels(factor(um[sid_tr, "Condition"]))
  Xte <- t(batch_res$batch_corrected[genes_val, sid_te, drop = FALSE])
  yte <- factor(um[sid_te, "Condition"], levels = levels(ytr))

  keep_tr <- stats::complete.cases(Xtr) & !is.na(ytr)
  Xtr <- Xtr[keep_tr, , drop = FALSE]
  ytr <- droplevels(ytr[keep_tr])

  if (ncol(Xtr) > 0L) {
    vr <- apply(Xtr, 2L, function(z) stats::var(z, na.rm = TRUE))
    ok <- is.finite(vr) & vr > 0
    Xtr <- Xtr[, ok, drop = FALSE]
    Xte <- Xte[, colnames(Xtr), drop = FALSE]
  }

  keep_te <- stats::complete.cases(Xte) & !is.na(yte)
  Xte <- Xte[keep_te, , drop = FALSE]
  yte <- droplevels(yte[keep_te])

  if (ncol(Xtr) >= 1L && nrow(Xtr) >= 3L && length(unique(ytr)) >= 2L && nrow(Xte) >= 1L) {
    fit_ext <- tryCatch(
      randomForest::randomForest(x = Xtr, y = ytr, ntree = 150L),
      error = function(e) NULL
    )
    if (!is.null(fit_ext)) {
      pr <- tryCatch(predict(fit_ext, Xte, type = "prob"), error = function(e) NULL)
      if (is.matrix(pr) && "Disease" %in% colnames(pr) && length(unique(yte)) >= 2L) {
        roc_te <- tryCatch(
          pROC::roc(yte, pr[, "Disease"], quiet = TRUE, levels = levels(ytr)),
          error = function(e) NULL
        )
        if (!is.null(roc_te)) {
          val_auc <- as.numeric(pROC::auc(roc_te))
          val_txt <- sprintf(
            "Held-out dataset D2: AUC = %.3f (train D1 / test D2, %d genes).",
            val_auc,
            ncol(Xtr)
          )
        }
      }
    }
  }
}

val_txt
```

# GSEA (gene-set enrichment analysis)

**Step 14 (GSEA)** ranks **all genes** by a differential statistic (here **limma logFC**) and tests whether curated sets are enriched at the **top or bottom** of the ranking (`clusterProfiler::gseGO()`). We show **GO BP**; the app can combine this with **MsigDB** collections.

```{r gsea-gsego}
gene_list <- de_df$logFC
names(gene_list) <- de_df$Gene
gene_list <- sort(gene_list, decreasing = TRUE)

gse_bp <- tryCatch(
  clusterProfiler::gseGO(
    geneList = gene_list,
    OrgDb = org.Hs.eg.db::org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    minGSSize = 10,
    maxGSSize = 500,
    pvalueCutoff = 0.5,
    eps = 0
  ),
  error = function(e) NULL
)

gse_df <- if (!is.null(gse_bp)) as.data.frame(gse_bp) else data.frame()
nrow(gse_df)
if (nrow(gse_df) > 0L) {
  head(gse_df[, intersect(c("Description", "NES", "pvalue", "p.adjust"), names(gse_df))], 5)
}
```

```{r gsea-kegg-optional}
## KEGG GSEA expects Entrez-ranked scores; map symbols from the same ranked list as gseGO.
gse_kegg <- NULL
gse_k_df <- data.frame()
if (length(gene_list) >= 10L) {
  kegg_map <- tryCatch(
    clusterProfiler::bitr(
      names(gene_list),
      fromType = "SYMBOL",
      toType = "ENTREZID",
      OrgDb = org.Hs.eg.db::org.Hs.eg.db
    ),
    error = function(e) data.frame()
  )
  if (nrow(kegg_map) > 5L) {
    sym_scores <- gene_list[kegg_map$SYMBOL]
    ez <- kegg_map$ENTREZID
    okm <- !is.na(sym_scores) & !is.na(ez)
    gl_entrez <- tapply(sym_scores[okm], ez[okm], mean, simplify = TRUE)
    gl_entrez <- sort(gl_entrez, decreasing = TRUE)
    gse_kegg <- tryCatch(
      clusterProfiler::gseKEGG(
        geneList = gl_entrez,
        organism = "hsa",
        minGSSize = 10,
        maxGSSize = 500,
        pvalueCutoff = 0.5,
        eps = 0
      ),
      error = function(e) NULL
    )
    gse_k_df <- if (!is.null(gse_kegg)) as.data.frame(gse_kegg) else data.frame()
  }
}
nrow(gse_k_df)
if (nrow(gse_k_df) > 0L) {
  head(
    gse_k_df[, intersect(c("ID", "Description", "NES", "p.adjust"), names(gse_k_df))],
    min(8L, nrow(gse_k_df))
  )
}
```

# MsigDB / Hallmark gene sets (GSEA inputs)

**Step 14** and pathway modules can use **Molecular Signatures Database** collections. The **`msigdbr`** package supplies tidy tables (no extra registration). Below we load **Hallmark (H)** gene sets for **Homo sapiens**—a typical input for ranked-list GSEA alongside `gseGO` / `gseKEGG`.

```{r msigdbr-hallmark}
hmk <- msigdbr(species = "Homo sapiens", category = "H")
n_sets <- length(unique(hmk$gs_name))
n_sets
head(hmk[, c("gs_name", "gene_symbol", "entrez_gene")], 10)
```

# Interactive Shiny application

`runGExPipe()` returns a **Shiny app object** (it does not call `runApp()` inside the package). Use a **fixed port** (e.g. `3838`) or `launch.browser = TRUE` so the browser URL is valid.

```{r app-object}
## Faster vignette check: defer heavy namespace attach (same pattern as other vignettes).
options(
  shiny.testmode = TRUE,
  gexpipe.minimal_attach_in_testmode = TRUE,
  gexpipe.wgcna_threads = 0L
)
app <- runGExPipe(launch.browser = FALSE, port = 3838L)
inherits(app, "shiny.appobj")
```

```{r launch-interactive, eval=FALSE}
if (interactive()) {
  app <- GExPipe::runGExPipe(launch.browser = TRUE, port = 3838)
  shiny::runApp(app, port = 3838)
}
```

# Further reading

- `vignettes/gexpipe-workflow.Rmd` — full runnable pipeline narrative and testing pointers.
- `vignettes/gexpipe-shiny-testing-ci.Rmd` — **shinytest2** smoke tests and CI notes.

# Optional GEO workflow (network-dependent)

Not evaluated during package checks; run locally when **GEOquery** access is available.

```{r optional-geo, eval=FALSE}
work_dir <- getwd()
gexp_prepare_download_dirs(work_dir, has_micro = TRUE, has_rna = FALSE)
gse_id <- "GSE10000"
micro_dir <- file.path(work_dir, "micro_data")
dl <- gexp_download_one_microarray_gse(gse_id, micro_dir)
dl$ok
```

# Session Info

```{r sessioninfo}
sessionInfo()
```

# References
