---
title: "GExPipe workflow: runnable R functions and Shiny entrypoint"
author: "GExPipe Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GExPipe workflow: runnable R functions and Shiny entrypoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
set.seed(123)
```

## Overview

`GExPipe` provides both:

- exported, testable R functions for core analysis steps, and
- a Shiny app object returned by `runGExPipe()`.

This vignette runs a complete lightweight path on bundled example data:

1. read expression + metadata from `inst/extdata`,
2. normalize and intersect genes,
3. batch-correct,
4. run differential expression,
5. run QC helper functions,
6. prepare WGCNA-ready input,
7. construct the Shiny app object.

The examples are designed to run during package checks.

## Setup

```{r setup-library}
library(GExPipe)
```

## Read bundled example data

```{r read-extdata}
vignette_extdata_file <- function(name) {
  p <- system.file("extdata", name, package = "GExPipe")
  if (nzchar(p) && file.exists(p)) return(p)

  candidates <- c(
    file.path(getwd(), "..", "inst", "extdata", name),
    file.path(getwd(), "..", "..", "inst", "extdata", name),
    file.path(getwd(), "inst", "extdata", name)
  )
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0L) {
    stop("Could not find bundled vignette file: ", name)
  }
  normalizePath(hits[1], winslash = "/", mustWork = TRUE)
}

expr_path <- vignette_extdata_file("vignette_expression.csv")
meta_path <- vignette_extdata_file("vignette_sample_metadata.csv")

tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- tab[[1]]
colnames(expr) <- gsub("^X", "", colnames(expr))

rownames(meta) <- meta$SampleID
meta <- meta[colnames(expr), , drop = FALSE]

dim(expr)
head(meta)
```

## Optional: parse GEO ID inputs

```{r parse-gse-inputs}
parsed <- gexp_parse_gse_inputs(
  analysis_type = "merged",
  rnaseq_gses = "GSE50760",
  microarray_gses = "GSE89076",
  dataset_mode = "multi"
)
parsed$rnaseq_ids
parsed$micro_ids
```

## Normalize and intersect common genes

```{r normalize-intersect}
cols_d1 <- rownames(meta)[meta$Dataset == "D1"]
cols_d2 <- rownames(meta)[meta$Dataset == "D2"]

micro_expr_list <- list(
  D1 = expr[, cols_d1, drop = FALSE],
  D2 = expr[, cols_d2, drop = FALSE]
)
rna_counts_list <- list()

norm <- gexp_normalize_and_intersect(
  micro_expr_list = micro_expr_list,
  rna_counts_list = rna_counts_list,
  micro_norm_method = "quantile",
  rnaseq_norm_method = "TMM",
  de_method = "limma"
)

dim(norm$combined_expr)
length(norm$common_genes)
```

```{r prepare-metadata}
um <- norm$unified_metadata
um$Condition <- meta[um$SampleID, "Condition"]
um$Dataset <- meta[um$SampleID, "Dataset"]
table(um$Dataset, um$Condition)
```

## Batch correction

```{r batch-correct}
batch_res <- gexp_batch_correct(
  expr = norm$combined_expr,
  metadata = um[, c("Dataset", "Condition")],
  variance_percentile = 25,
  method = "limma"
)

dim(batch_res$expr_filtered)
dim(batch_res$batch_corrected)
```

## Differential expression (limma)

```{r run-de}
de_res <- gexp_run_de(
  expr = batch_res$batch_corrected,
  metadata = um[, c("Dataset", "Condition")],
  method = "limma",
  logfc_cutoff = 0.25,
  padj_cutoff = 0.10
)

head(de_res$de_results[, c("Gene", "logFC", "adj.P.Val", "Significance")])
```

## QC helper functions

```{r qc-helpers}
qc_out <- gexp_qc_detect_outliers(norm$combined_expr, top_n = 100)
length(qc_out$all_outliers)

qc_summary <- gexp_qc_gene_overlap_summary(
  all_genes_list = list(D1 = rownames(micro_expr_list$D1), D2 = rownames(micro_expr_list$D2)),
  common_genes = norm$common_genes
)
qc_summary
```

## WGCNA preparation

```{r wgcna-prepare}
wgcna_prep <- gexp_wgcna_prepare(
  expr = batch_res$batch_corrected,
  metadata = um,
  gene_mode = "top_variable",
  top_genes = min(100L, nrow(batch_res$batch_corrected)),
  min_samples_frac = 0.5
)

dim(wgcna_prep$datExpr)
head(wgcna_prep$gene_variance_table, 3)
```

## Shiny app object

The package function returns a Shiny app object; launch only in interactive sessions.

```{r app-object}
app <- runGExPipe(launch.browser = FALSE)
inherits(app, "shiny.appobj")
```

```{r launch-app-interactive, eval=FALSE}
if (interactive()) {
  shiny::runApp(app)
}
```

## Shiny testing and CI notes

For reviewer-facing Shiny test execution details (including `shinytest2` smoke tests and non-interactive CI behavior), see:

- `vignettes/gexpipe-shiny-testing-ci.Rmd`

## If something is missing: manual fill-in instructions

If your local branch still has incomplete exported functions or data objects, use this checklist:

1. Confirm exports are visible:
   - run `getNamespaceExports("GExPipe")`,
   - ensure functions used in this vignette are listed.
2. If a function name changed, replace it in the corresponding chunk and keep one runnable call.
3. If bundled files are missing:
   - add them to `inst/extdata/`,
   - verify with `system.file("extdata", "<name>", package = "GExPipe")`.
4. If a step is too heavy for check, keep demonstration code in a chunk with `eval = FALSE` and add a short explanation.
5. Rebuild docs and run checks:
   - `devtools::document()`
   - `devtools::build_vignettes()`
   - `devtools::test()`
   - `devtools::check()`

## Optional GEO workflow (interactive / network-dependent)

These commands are examples for local interactive use and are not run during checks.

```{r optional-geo-workflow, eval=FALSE}
work_dir <- getwd()
gexp_prepare_download_dirs(work_dir, has_micro = TRUE, has_rna = FALSE)

gse_id <- "GSE10000"
micro_dir <- file.path(work_dir, "micro_data")
dl <- gexp_download_one_microarray_gse(gse_id, micro_dir)
dl$ok
```

## Session information

```{r session-info}
sessionInfo()
```
