---
title: "GExPipe: A Pipeline for Gene Expression Analysis"
author:
  - name: Safa Rafique, Dr. Naeem Mahmood Ashraf, Prof. Dr. Farooq sabir
    affiliation: University of the Punjab

date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to GExPipe}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`GExPipe` is a Shiny-first workflow for bulk RNA-seq and microarray
analysis. The package was created to provide a single, guided pipeline
that connects data retrieval, quality control, normalization,
differential expression, network analysis, and downstream interpretation
without requiring users to manually stitch many separate tools together.

Within Bioconductor, `GExPipe` builds on established infrastructure such
as `r BiocStyle::Biocpkg("GEOquery")`, `r BiocStyle::Biocpkg("limma")`,
`r BiocStyle::Biocpkg("DESeq2")`, `r BiocStyle::Biocpkg("edgeR")`,
`r BiocStyle::Biocpkg("WGCNA")`, and
`r BiocStyle::Biocpkg("clusterProfiler")`. For visualization it uses
`r BiocStyle::CRANpkg("ggplot2")` and
`r BiocStyle::CRANpkg("pheatmap")`. Compared with piecemeal scripts, the
app emphasizes usability and reproducibility through a fixed
step-by-step interface and integrated summaries.

This vignette is user-oriented: it explains how to install, launch, and
run a minimal executable workflow using package-bundled example data.

# Installation

Install `GExPipe` with `BiocManager` once the package has been accepted
on Bioconductor:

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GExPipe")
```

Until then, install directly from GitHub:

```{r install-github, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Core Bioconductor dependencies
BiocManager::install(c(
    "DESeq2", "GEOquery", "limma", "edgeR", "Biobase", "AnnotationDbi",
    "org.Hs.eg.db", "clusterProfiler", "enrichplot", "STRINGdb", "sva",
    "WGCNA", "dynamicTreeCut", "msigdbr"
), update = FALSE, ask = FALSE)

if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")

remotes::install_github("safarafique/GExPipe")
```

# Launch the Shiny app

`runGExPipe()` returns a Shiny app object. Pass it to `shiny::runApp()`
to start the server. The function does **not** launch the app by itself
— this follows the Bioconductor convention for Shiny packages.

> **Note:** The block below is not evaluated during vignette build
> because Shiny apps require an interactive session. Copy it into an R
> console to launch the app.

```{r launch-app, eval=FALSE}
app <- GExPipe::runGExPipe(launch.browser = TRUE, port = 3838L)
shiny::runApp(app, port = 3838L)
```

# Workflow Example (evaluated)

This section demonstrates three core tasks with bundled data:

-   loading expression and metadata tables,
-   running basic processing suitable for downstream analysis,
-   visualizing results (PCA and expression heatmap).

## Load package and example data

```{r load-data}
library(GExPipe)
library(ggplot2)

expr_path <- system.file("extdata", "vignette_expression.csv", package = "GExPipe")
meta_path <- system.file("extdata", "vignette_sample_metadata.csv", package = "GExPipe")

stopifnot(nzchar(expr_path), file.exists(expr_path))
stopifnot(nzchar(meta_path), file.exists(meta_path))

expr_tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta_tab <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(expr_tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- expr_tab[[1]]

rownames(meta_tab) <- meta_tab$SampleID
meta_tab <- meta_tab[colnames(expr), , drop = FALSE]

dim(expr)
head(meta_tab)
```

## Process data with core pipeline helpers

`gexp_normalize_and_intersect()` accepts a named list of per-dataset
expression matrices (`micro_expr_list` for microarray / pre-normalized
data, `rna_counts_list` for raw RNA-seq counts) and returns a list with:

-   **`combined_expr`** — a single genes × samples matrix after
    within-dataset normalization and global quantile normalization
    across all datasets.
-   **`common_genes`** — the character vector of genes present in every
    dataset.
-   **`combined_expr_before_global`** — the combined matrix before the
    final global quantile step (useful for QC comparisons).

```{r process-data}
if (!"Dataset" %in% colnames(meta_tab)) {
  meta_tab$Dataset <- "D1"
}

# Build a named list: one matrix per dataset (genes × samples)
datasets <- unique(meta_tab$Dataset)
micro_expr_list <- setNames(
  lapply(datasets, function(ds) {
    sample_ids <- rownames(meta_tab)[meta_tab$Dataset == ds]
    expr[, sample_ids, drop = FALSE]
  }),
  datasets
)

norm_res <- gexp_normalize_and_intersect(
  micro_expr_list    = micro_expr_list,
  rna_counts_list    = list(),
  micro_norm_method  = "quantile",
  rnaseq_norm_method = "TMM",
  de_method          = "limma"
)

cat("Combined matrix dimensions:", dim(norm_res$combined_expr), "\n")
cat("Common genes across datasets:", length(norm_res$common_genes), "\n")
```

## Visualize processed data

### PCA plot

```{r pca-plot}
# Remove near-zero-variance genes before scaling to avoid NaN in prcomp
expr_for_pca <- norm_res$combined_expr
keep_var <- apply(expr_for_pca, 1, var, na.rm = TRUE) > 1e-8
pca <- prcomp(t(expr_for_pca[keep_var, ]), center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  Sample  = rownames(pca$x),
  PC1     = pca$x[, 1],
  PC2     = pca$x[, 2],
  Dataset = meta_tab[rownames(pca$x), "Dataset", drop = TRUE],
  stringsAsFactors = FALSE
)

var_pct <- round(summary(pca)$importance[2, 1:2] * 100, 1)

ggplot(pca_df, aes(PC1, PC2, color = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title    = "PCA of bundled example data",
    subtitle = "After normalization and gene intersection",
    x        = paste0("PC1 (", var_pct[1], "% variance)"),
    y        = paste0("PC2 (", var_pct[2], "% variance)")
  )
```

### Expression heatmap

```{r heatmap-plot}
top_var <- order(apply(norm_res$combined_expr, 1, var, na.rm = TRUE),
                 decreasing = TRUE)
top_var <- head(top_var, min(30L, length(top_var)))

if (length(top_var) > 1L) {
  pheatmap::pheatmap(
    norm_res$combined_expr[top_var, , drop = FALSE],
    scale         = "row",
    show_rownames = length(top_var) <= 30L,
    main          = "Top variable genes (bundled example)"
  )
}
```

# App walkthrough checklist

For real analyses in the Shiny app, use this sequence:

1.  **Step 1 — Download:** select platform (RNA-seq / Microarray /
    Merged) and GEO IDs, then click *Start Processing*.
2.  **Step 2 — QC:** review boxplots, density plots, and outlier flags;
    remove samples if needed.
3.  **Steps 3–5 — Normalize / Groups / Batch:** choose normalization
    method, annotate conditions, and apply ComBat or limma batch
    correction.
4.  **Steps 6–8 — DE / WGCNA / Common genes:** run differential
    expression, build co-expression modules, and intersect DEGs with
    WGCNA hub genes.
5.  **Steps 9–15 — Enrichment / PPI / ML / Validation / ROC / Nomogram /
    GSEA / Summary:** downstream interpretation, machine-learning
    feature selection, and final PDF report generation.

# App screenshots

If screenshots are placed under `inst/extdata/vignette-screenshots/`,
they are rendered below.

```{r screenshots, echo=FALSE, results='asis'}
shot_dir <- system.file("extdata", "vignette-screenshots", package = "GExPipe")

shot_meta <- list(
  list(file = "step1_download.png",  label = "Step 1 — Download Data"),
  list(file = "step2_qc.png",        label = "Step 2 — Quality Control"),
  list(file = "step6_de.png",        label = "Step 6 — Differential Expression"),
  list(file = "step9_ppi.png",       label = "Step 9 — PPI Network"),
  list(file = "step15_summary.png",  label = "Step 15 — Results Summary")
)

if (nzchar(shot_dir)) {
  for (item in shot_meta) {
    p <- file.path(shot_dir, item$file)
    if (file.exists(p)) {
      cat("\n\n## ", item$label, "\n\n", sep = "")
      print(knitr::include_graphics(p))
    }
  }
}
```

# Session Information

```{r sessionInfo}
sessionInfo()
```
