---
title: "GExPipe Shiny testing and CI reproducibility"
author: "GExPipe Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GExPipe Shiny testing and CI reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Why this vignette exists

Bioconductor review expects Shiny packages to be tested in addition to standard unit tests.\
This vignette documents how `GExPipe` is exercised in:

-   regular testthat runs,
-   non-interactive CI-like sessions, and
-   local manual smoke checks.

## App object contract

The package exposes `runGExPipe()` as the Shiny entrypoint. In vignette/check context we keep this check side-effect free (no port binding).

```{r app-contract}
library(GExPipe)

exists("runGExPipe", where = asNamespace("GExPipe"), inherits = FALSE)
"launch.browser" %in% names(formals(GExPipe::runGExPipe))
```

Suggested testing packages (Bioconductor [Shiny testing section](https://contributions.bioconductor.org/shiny.html#testing)):

```{r testing-suggests-available}
data.frame(
  package = c("testthat", "shinytest2", "chromote"),
  available = vapply(
    c("testthat", "shinytest2", "chromote"),
    requireNamespace,
    logical(1L),
    quietly = TRUE
  ),
  row.names = NULL
)
```

Launch only in interactive use. Keep **`app <- runGExPipe(...)` and `runApp(app)` inside one `if (interactive()) { ... }` block** (as below). That way `app` always exists before `runApp`, and if someone knits with `eval=TRUE` by mistake, non-interactive builds skip the whole block (no server, no stray `runApp(app)` after a failed assignment).

**Important:** `shiny::runApp()` **blocks** the R session: the console stays busy until you **Stop** the app (RStudio red Stop icon). That is normal—not a hang.

**How to tell “working” vs “broken”:**

| What you see | Meaning |
|----|----|
| `GExPipe: building Shiny UI...` then (pause) then `GExPipe: app object built` | `runGExPipe()` finished; pause can be **minutes** on first run. |
| After `runApp(app)`: `GExPipe: Shiny onStart...` then **`Listening on http://127.0.0.1:PORT`** | Server is up — open that URL in a browser (`launch.browser = FALSE` does not open it for you). |
| **No** `>` prompt while the app runs | **Expected** — the session is serving the app, not stuck. |
| **No** `Listening on...` after **10+ minutes** | Unusual — interrupt (Stop), update packages, try again; avoid `inst/shinyapp` Run App if you see `global.R` “Loading packages” text (slower path). |

This chunk stays `eval=FALSE` for knitting; to try the app, run **the entire chunk** in the Console, not only the `runApp` lines.

```{r launch-interactive, eval=FALSE}
# Use a fixed port (e.g. 3838) or launch.browser = TRUE — port = 0 is not a valid browser URL.
if (interactive()) {
  app <- GExPipe::runGExPipe(launch.browser = TRUE, port = 3838)
  shiny::runApp(app, port = 3838)
}
```

## Automated Shiny smoke test

The package includes a `shinytest2` smoke test (see `tests/testthat/test-shiny-integration.R`) that:

1.  builds an app from package source,
2.  starts an `AppDriver`,
3.  confirms successful app startup.

Run only that test (from package root):

```{r run-shinytest-only, eval=FALSE}
testthat::test_local(filter = "shiny-integration")
```

Equivalent:

```{r run-shinytest-only-devtools, eval=FALSE}
devtools::test(filter = "shiny-integration")
```

Run all tests:

```{r run-all-tests, eval=FALSE}
testthat::test_local()
```

```{r run-all-tests-devtools, eval=FALSE}
devtools::test()
```

## Non-interactive CI behavior

The Shiny smoke test is designed to run in non-interactive sessions (for CI/check contexts), with a controlled opt-out:

-   if `CI=true` and `GEXPIPE_SKIP_SHINYTEST2=true`, the test is skipped intentionally,
-   otherwise it attempts to run and reports startup issues explicitly.

Example CI-style run:

```{r ci-style-run, eval=FALSE}
Sys.setenv(CI = "true")
Sys.unsetenv("GEXPIPE_SKIP_SHINYTEST2")
devtools::test(filter = "shiny-integration")
Sys.unsetenv("CI")
```

Optional opt-out:

```{r ci-opt-out, eval=FALSE}
Sys.setenv(CI = "true", GEXPIPE_SKIP_SHINYTEST2 = "true")
devtools::test(filter = "shiny-integration")
Sys.unsetenv("CI")
Sys.unsetenv("GEXPIPE_SKIP_SHINYTEST2")
```

## Manual reviewer checklist

Run the following in a fresh R session from package root:

```{r manual-reviewer-checklist, eval=FALSE}
pkgload::load_all(".", quiet = TRUE, export_all = FALSE)
inherits(GExPipe::runGExPipe(FALSE), "shiny.appobj")

testthat::test_local(filter = "shiny-integration")
testthat::test_local()

tarball <- devtools::build(path = tempdir())
rcmdcheck::rcmdcheck(tarball, args = "--no-manual", error_on = "warning")
BiocCheck::BiocCheck(tarball, `quit-with-status` = FALSE)
```

## Troubleshooting

-   **Chromium/Chrome not available**: install/update browser backend used by `chromote`.
-   **App startup timeout** (`did not become stable`): shinytest2 blocks until `window.shinytest2.ready` (Shiny session initialized). The app defers heavy namespace attachment when **`options(shiny.testmode = TRUE)`** and **`gexpipe.minimal_attach_in_testmode`** is true (default): only **shiny / shinydashboard / shinyjs / DT** attach before the first response; the rest run after **`session$onFlushed()`**. The smoke test child uses **`library(GExPipe)`** when installed (`R CMD check`), or **`pkgload::load_all()`** from source (slower; set **`GEXPIPE_SHINYTEST2_USE_PKGLOAD=1`** to force it). Increase **`GEXPIPE_SHINYTEST2_MS`** if the child is slow (default **600000** ms). `AppDriver(..., wait = FALSE)` skips the 200 ms idle wait. With `shiny.testmode = TRUE`, the 5s network poll is off.
-   **Environment-only failures**: re-run with explicit `CI` and `GEXPIPE_SKIP_SHINYTEST2` settings to verify gating behavior.

## Session information

```{r session-info}
sessionInfo()
```
