---
title: "GExPipe: An Integrated Pipeline for Gene Expression Analysis"
author:
  - name: Safa Rafique, Naeem Mahmood Ashraf, Farooq Sabir
    affiliation: University of the Punjab
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to GExPipe}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

# Introduction

`GExPipe` is a comprehensive, Shiny-based workflow for the analysis of
bulk RNA-seq and microarray data. By integrating a suite of Bioconductor
tools into a single guided interface, it bridges the gap between raw
GEO datasets and actionable biological insight — without requiring users
to manually connect separate tools.

The package builds on `r BiocStyle::Biocpkg("GEOquery")` for data
retrieval, `r BiocStyle::Biocpkg("limma")`,
`r BiocStyle::Biocpkg("DESeq2")`, and
`r BiocStyle::Biocpkg("edgeR")` for differential expression,
`r BiocStyle::Biocpkg("WGCNA")` for co-expression networks, and
`r BiocStyle::Biocpkg("clusterProfiler")` for functional enrichment.

## Analysis Pipeline Overview

`GExPipe` organises the analysis into 15 sequential steps across four
phases. Each step must be completed before the next becomes available.

| Phase | Steps | What happens |
|---|---|---|
| **Data preparation** | 1 – 5 | Download GEO data → Quality control → Normalization → Group assignment → Batch correction |
| **Gene discovery** | 6 – 8 | Differential expression → WGCNA co-expression network → DEG ∩ hub gene intersection |
| **Candidate refinement** | 9 – 12 | PPI network → ML feature selection → Validation → ROC analysis |
| **Clinical translation** | 13 – 15 | Nomogram → GSEA pathway enrichment → Summary report (PDF) |

# Installation

## From Bioconductor (Recommended — after acceptance)

Once accepted into Bioconductor, install the stable release with two
commands. `BiocManager` automatically selects the correct Bioconductor
release for your R version (3.21 for R 4.5, 3.22 for R 4.6, etc.):

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GExPipe")
```

Then launch the app:

```{r launch-bioc, eval=FALSE}
GExPipe::runGExPipe()
```

## From GitHub (Current — available now)

Install the package from GitHub and launch the app with three commands.
Works on **Windows, macOS, and Linux** without any manual dependency
management:

```{r install-github, eval=FALSE}
options(timeout = 3600)   # 1 hour — needed for slow connections

# Step 1: Install BiocManager and set repos so remotes can find
#         Bioconductor packages (org.Hs.eg.db, DESeq2, etc.) automatically
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
options(repos = BiocManager::repositories())

# Step 2: Install GExPipe from GitHub (all 54 dependencies installed automatically)
if (!requireNamespace("remotes", quietly = TRUE))
    install.packages("remotes")
remotes::install_github("safarafique/GExPipe", dependencies = TRUE)

GExPipe::runGExPipe()
```

The app handles all 54 dependencies automatically on first launch:

- Creates an **isolated GExPipe library** on your system
- Downloads and installs all required packages into it
- Keeps your main R library untouched
- On every subsequent run, opens in seconds

### Isolated library location

All GExPipe packages are installed into a dedicated folder, completely
separate from your main R library:

| Operating System | Path |
|---|---|
| **Windows** | `C:\Users\<name>\AppData\Local\GExPipe\<R-ver>\` |
| **macOS** | `~/Library/Application Support/GExPipe/<R-ver>/` |
| **Linux** | `~/.local/share/GExPipe/<R-ver>/` |

To find the exact path on your system:

```{r lib-path, eval=FALSE}
rv <- paste0(R.Version()$major, ".", sub("\\..*", "", R.Version()$minor))
file.path(Sys.getenv("LOCALAPPDATA"), "GExPipe", rv)   # Windows
# ~/Library/Application Support/GExPipe/<rv>            # macOS
# ~/.local/share/GExPipe/<rv>                           # Linux
```

> **Rtools (Windows only):** If any dependency requires compilation,
> install [Rtools45](https://cran.r-project.org/bin/windows/Rtools/)
> once. R will then compile source packages automatically.
>
> **First run time:** Downloading and installing all 54 packages takes
> approximately 10–40 minutes depending on internet speed. Every
> subsequent run opens in seconds.

# Quick Start

## Launch the App

After installation, start the application with a single command:

```{r launch, eval=FALSE}
GExPipe::runGExPipe()
```

The app opens in your default browser at **Step 1 — Download Data**.

On **first launch** the startup sequence runs automatically:

```
======================================================================
  GExPipe — loading 54 packages...
  [ 1/54] shiny             ... ✓ 1.13.0   Ready
  [ 2/54] shinydashboard    ... ✓ 0.7.3    Ready
  ...
  [54/54] xgboost           ... ✓ 3.2.1.1  Ready
  ===============================================================
  Loaded  : 54 / 54 packages
  Status  : All packages ready — opening app...
  ===============================================================
```

On **subsequent launches** all packages are already in the isolated
library — the app opens in seconds with no install step.

## Launch in the Cloud (Google Colab)

For users without a local R environment, `GExPipe` can be run in
Google Colab at no cost. Open a new notebook, change the runtime to R
(**Runtime → Change runtime type → R**), then run the two cells below.

**Cell 1 — Start a public tunnel** (free token from
<https://ngrok.com>):

```{r colab-tunnel, eval=FALSE}
NGROK_TOKEN <- "PASTE_YOUR_TOKEN_HERE"

system(paste0(
    "wget -q https://bin.equinox.io/c/bNyj1mQVY4c/",
    "ngrok-v3-stable-linux-amd64.tgz -O ngrok.tgz",
    " && tar xzf ngrok.tgz"
), wait = TRUE)
system(paste0("./ngrok config add-authtoken ", NGROK_TOKEN), wait = TRUE)
system("./ngrok http 3838 > /dev/null 2>&1 &", wait = FALSE)
Sys.sleep(3)

url <- system(
    paste0(
        "curl -s http://localhost:4040/api/tunnels | ",
        "python3 -c \"import sys,json; t=json.load(sys.stdin)['tunnels']; ",
        "print(t[0]['public_url'] if t else 'ngrok not ready')\""
    ),
    intern = TRUE
)
cat("Open this URL in your browser:", url, "\n")
```

**Cell 2 — Run the app** (the cell blocks while the app is running —
this is expected):

```{r colab-run, eval=FALSE}
options(timeout = 3600)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager", repos = "https://cloud.r-project.org",
                     quiet = TRUE)

BiocManager::install("safarafique/GExPipe",
                     dependencies = TRUE, ask = FALSE)

GExPipe::runGExPipe(host = "0.0.0.0", port = 3838L,
                    launch.browser = FALSE)
```

# Workflow Example

This section demonstrates the core programmatic API using bundled
example data: loading an expression matrix and sample metadata,
applying normalization and cross-dataset gene intersection, and
visualizing results with PCA and a heatmap.

## Load package and example data

```{r load-data}
library(GExPipe)
library(ggplot2)

expr_path <- system.file("extdata", "vignette_expression.csv",
                          package = "GExPipe")
meta_path <- system.file("extdata", "vignette_sample_metadata.csv",
                          package = "GExPipe")

stopifnot(nzchar(expr_path), file.exists(expr_path))
stopifnot(nzchar(meta_path), file.exists(meta_path))

expr_tab <- read.csv(expr_path, check.names = FALSE,
                     stringsAsFactors = FALSE)
meta_tab <- read.csv(meta_path, check.names = FALSE,
                     stringsAsFactors = FALSE)

expr <- as.matrix(expr_tab[, -1L, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- expr_tab[[1L]]

rownames(meta_tab) <- meta_tab$SampleID
meta_tab <- meta_tab[colnames(expr), , drop = FALSE]

dim(expr)
head(meta_tab)
```

## Normalize and intersect datasets

`gexp_normalize_and_intersect()` accepts a named list of expression
matrices, applies within-dataset normalization, and returns a single
genes × samples matrix restricted to genes present across all datasets.

```{r normalize}
if (!"Dataset" %in% colnames(meta_tab))
    meta_tab$Dataset <- "D1"

datasets <- unique(meta_tab$Dataset)
micro_expr_list <- setNames(
    lapply(datasets, function(ds) {
        ids <- rownames(meta_tab)[meta_tab$Dataset == ds]
        expr[, ids, drop = FALSE]
    }),
    datasets
)

norm_res <- gexp_normalize_and_intersect(
    micro_expr_list    = micro_expr_list,
    rna_counts_list    = list(),
    micro_norm_method  = "quantile",
    rnaseq_norm_method = "TMM",
    de_method          = "limma"
)

cat("Combined matrix :", paste(dim(norm_res$combined_expr), collapse = " x "), "\n")
cat("Common genes    :", length(norm_res$common_genes), "\n")
```

## Principal component analysis

```{r pca-plot}
expr_pca <- norm_res$combined_expr
keep     <- apply(expr_pca, 1L, var, na.rm = TRUE) > 1e-8
pca      <- prcomp(t(expr_pca[keep, ]), center = TRUE, scale. = TRUE)

pca_df <- data.frame(
    PC1     = pca$x[, 1L],
    PC2     = pca$x[, 2L],
    Dataset = meta_tab[rownames(pca$x), "Dataset"],
    stringsAsFactors = FALSE
)
var_pct <- round(summary(pca)$importance[2L, 1L:2L] * 100, 1)

ggplot(pca_df, aes(PC1, PC2, colour = Dataset)) +
    geom_point(size = 3) +
    theme_minimal() +
    labs(
        title    = "PCA — bundled example data",
        subtitle = "After quantile normalization and gene intersection",
        x        = paste0("PC1 (", var_pct[1L], "% variance)"),
        y        = paste0("PC2 (", var_pct[2L], "% variance)")
    )
```

## Expression heatmap

```{r heatmap-plot}
top_var <- head(
    order(apply(norm_res$combined_expr, 1L, var, na.rm = TRUE),
          decreasing = TRUE),
    min(30L, nrow(norm_res$combined_expr))
)

if (length(top_var) > 1L) {
    pheatmap::pheatmap(
        norm_res$combined_expr[top_var, , drop = FALSE],
        scale         = "row",
        show_rownames = length(top_var) <= 30L,
        main          = "Top 30 most variable genes — bundled example data"
    )
}
```

# Application Walkthrough — 15-Step Pipeline

The Shiny interface enforces a linear pipeline; each step is locked
until its predecessor is complete.

## Step 1 — Download Data

Retrieves one or more GEO datasets and assembles a combined expression
matrix. The user specifies a platform type (Microarray, RNA-seq, or
Merged) and one or more GSE accession numbers
(e.g. `GSE89076, GSE50760`). Series-matrix files are downloaded via
`r BiocStyle::Biocpkg("GEOquery")`, probe or Ensembl identifiers are
mapped to HGNC symbols, duplicate probes are averaged, and only genes
present across all datasets are retained.

## Step 2 — Quality Control

Identifies and removes outlier samples before normalization. Per-sample
boxplots and kernel-density estimates are rendered; statistical outliers
are flagged by Mahalanobis distance (*p* < 0.001 by default). A minimum
of three samples per group must remain after exclusion.

## Step 3 — Normalize

Scales expression values to be comparable across samples and datasets.
*Auto* mode applies quantile normalization to microarray data and TMM
to RNA-seq counts. *Manual* mode offers quantile, log~2~, VST, RLE,
and upper-quantile methods. Successful normalization is confirmed by
aligned sample medians in the post-normalization boxplot.

## Step 4 — Assign Groups

Maps each sample to *Normal*, *Disease*, or *Exclude* using a selected
metadata column. A summary panel confirms at least three samples in
each retained class before proceeding.

## Step 5 — Batch Correction

Removes systematic technical variation introduced by different
experimental batches or sequencing runs. Available methods are ComBat
(parametric empirical Bayes, `r BiocStyle::Biocpkg("sva")`) and
`removeBatchEffect` (`r BiocStyle::Biocpkg("limma")`). This step is
automatically skipped when only one dataset is loaded.

## Step 6 — Differential Expression

Identifies genes with statistically significant expression changes
between Normal and Disease. Default thresholds are FDR < 0.05 and
|log~2~FC| > 1. Available methods are limma (microarray and
log-normalized RNA-seq), limma-voom, DESeq2, and edgeR (count data).
Outputs include a volcano plot, a heatmap of the top DEGs, and a
downloadable results table.

## Step 7 — WGCNA Co-expression Network

Clusters genes into co-expression modules and correlates each module
eigengene with the disease trait using
`r BiocStyle::Biocpkg("WGCNA")`. The user selects the soft-threshold
power β where the scale-free topology fit R² ≥ 0.80, sets a minimum
module size (default 30), and a merge cut height for combining similar
modules (default 0.25). Outputs include the scale-free topology plot,
gene dendrogram, and module–trait correlation heatmap.

## Step 8 — Common Genes

Intersects the DEG list from Step 6 with hub genes from
trait-associated modules identified in Step 7. The intersection is
subjected to Gene Ontology and KEGG over-representation analysis via
`r BiocStyle::Biocpkg("clusterProfiler")`. Outputs include a Venn
diagram, GO and KEGG dot plots, and a downloadable candidate gene table.

## Step 9 — PPI Network

Maps the candidate gene set onto a protein–protein interaction network
from STRINGdb and ranks proteins by graph-theoretic centrality. The
user specifies a confidence score cutoff (default 400) and network
layout. Outputs include an interactive network plot and a hub-protein
table reporting degree, betweenness, and eigenvector centrality.

## Step 10 — Machine-Learning Feature Selection

Applies an ensemble of algorithms to identify the smallest gene panel
that maximally discriminates Normal from Disease samples. Algorithms:
LASSO, Elastic Net, Ridge, Random Forest, SVM-RFE, Boruta, sPLS-DA,
and XGBoost with SHAP importance scores. Genes selected by two or more
algorithms constitute the final candidate set.

## Step 11 — Validation

Confirms that the candidate gene set generalizes beyond the discovery
cohort. *External* validation downloads an independent GEO dataset
automatically; *internal* validation applies a stratified 70/30
train–test split. Classification accuracy ≥ 70% in the validation set
is the minimum acceptance criterion.

## Step 12 — ROC Analysis

Quantifies the individual diagnostic accuracy of each candidate gene.
A logistic regression model is fitted per gene; the ROC curve is
computed with `r BiocStyle::CRANpkg("pROC")` and the AUC is reported
with a 95% confidence interval. Genes with AUC ≥ 0.80 in both the
training and validation sets are retained.

## Step 13 — Nomogram

Constructs a clinically interpretable scoring model from the final gene
panel using `rms::lrm()`. Each gene is assigned a point scale whose
sum maps to a predicted disease probability. A calibration curve and
decision curve analysis (DCA) are generated to assess model reliability
and clinical utility.

## Step 14 — GSEA

Identifies coordinated biological pathway activity in samples
stratified by expression of each final gene. Gene sets tested include
Gene Ontology (BP, MF, CC), KEGG, and MSigDB collections (Hallmark,
C2, C5). Outputs include a ridge plot, individual enrichment plots, and
a results table with normalized enrichment scores (NES) and
leading-edge gene lists.

## Step 15 — Summary Report

Generates a structured narrative and a downloadable PDF covering the
complete analysis: dataset details, DEG counts, WGCNA module findings,
common gene intersection, PPI hub proteins, machine-learning features,
validation accuracy, ROC AUCs, and top enriched pathways. The report is
rendered via `rmarkdown`.

# Troubleshooting {#troubleshooting}

| Problem | Solution |
|---|---|
| **Locked DLL files (Windows)** | Restart R (`Ctrl+Shift+F10`) and run the installation command first, before loading any other packages. |
| **Missing C++ compiler** | Install [Rtools45](https://cran.r-project.org/bin/windows/Rtools/) (Windows only). |
| **Version conflicts** (`namespace X loaded but >= Y required`) | Use `pak::pkg_install("safarafique/GExPipe")` for smarter dependency resolution. |
| **Corrupt lazy-load database** | Reinstall with `BiocManager::install("safarafique/GExPipe", INSTALL_opts = "--no-staged-install")`. |
| **Corporate proxy / firewall** | Set `Sys.setenv(https_proxy = "http://proxy:port")` before installing. |
| **Offline environment** | Download the ZIP from GitHub and run `install.packages("GExPipe-main.zip", repos = NULL, type = "source")`. |

## Windows: Persistent DLL lock errors

In rare cases — typically after a failed package update — R holds a
lock on `curl.dll` that prevents all subsequent downloads. If
restarting R does not resolve the issue, the following block clears
stale lock directories and switches to the Windows-native HTTP stack:

```{r windows-fix, eval=FALSE}
# Switch to Windows-native HTTP (bypasses curl.dll entirely)
options(download.file.method = "wininet")

# Remove stale lock directories left by the failed install
unlink(
    list.files(.libPaths()[1], pattern = "^00LOCK-", full.names = TRUE),
    recursive = TRUE, force = TRUE
)

# Retry installation
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("safarafique/GExPipe", dependencies = TRUE)
```

After installation completes, restart R once more and launch normally
with `GExPipe::runGExPipe()`.

# Session Information

```{r session-info}
sessionInfo()
```
