---
title: "GExPipe Shiny testing and CI reproducibility"
author: "GExPipe Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GExPipe Shiny testing and CI reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Why this vignette exists

Bioconductor review expects Shiny packages to be tested in addition to standard unit tests.  
This vignette documents how `GExPipe` is exercised in:

- regular testthat runs,
- non-interactive CI-like sessions, and
- local manual smoke checks.

## App object contract

The package exposes `runGExPipe()` as the Shiny entrypoint. In vignette/check context we keep this check side-effect free (no port binding).

```{r app-contract}
library(GExPipe)

exists("runGExPipe", where = asNamespace("GExPipe"), inherits = FALSE)
"launch.browser" %in% names(formals(GExPipe::runGExPipe))
```

Launch only in interactive use:

```{r launch-interactive, eval=FALSE}
if (interactive()) {
  shiny::runApp(app)
}
```

## Automated Shiny smoke test

The package includes a `shinytest2` smoke test (see `tests/testthat/test-shiny-integration.R`) that:

1. builds an app from package source,
2. starts an `AppDriver`,
3. confirms successful app startup.

Run only that test:

```{r run-shinytest-only, eval=FALSE}
devtools::test(filter = "shiny-integration")
```

Run all tests:

```{r run-all-tests, eval=FALSE}
devtools::test()
```

## Non-interactive CI behavior

The Shiny smoke test is designed to run in non-interactive sessions (for CI/check contexts), with a controlled opt-out:

- if `CI=true` and `GEXPIPE_SKIP_SHINYTEST2=true`, the test is skipped intentionally,
- otherwise it attempts to run and reports startup issues explicitly.

Example CI-style run:

```{r ci-style-run, eval=FALSE}
Sys.setenv(CI = "true")
Sys.unsetenv("GEXPIPE_SKIP_SHINYTEST2")
devtools::test(filter = "shiny-integration")
Sys.unsetenv("CI")
```

Optional opt-out:

```{r ci-opt-out, eval=FALSE}
Sys.setenv(CI = "true", GEXPIPE_SKIP_SHINYTEST2 = "true")
devtools::test(filter = "shiny-integration")
Sys.unsetenv("CI")
Sys.unsetenv("GEXPIPE_SKIP_SHINYTEST2")
```

## Manual reviewer checklist

Run the following in a fresh R session from package root:

```{r manual-reviewer-checklist, eval=FALSE}
pkgload::load_all(".", quiet = TRUE, export_all = FALSE)
inherits(GExPipe::runGExPipe(FALSE), "shiny.appobj")

devtools::test(filter = "shiny-integration")
devtools::test()

tarball <- devtools::build(path = tempdir())
rcmdcheck::rcmdcheck(tarball, args = "--no-manual", error_on = "warning")
BiocCheck::BiocCheck(".", `quit-with-status` = FALSE)
```

## Troubleshooting

- **Chromium/Chrome not available**: install/update browser backend used by `chromote`.
- **App startup timeout**: inspect test logs and check duplicate IDs or long blocking operations at startup.
- **Environment-only failures**: re-run with explicit `CI` and `GEXPIPE_SKIP_SHINYTEST2` settings to verify gating behavior.

## Session information

```{r session-info}
sessionInfo()
```
