---
title: "GExPipe: A Pipeline for Gene Expression Analysis"
author:
  - name: Safa Rafique, Dr. Naeem Mahmood Ashraf, Prof. Dr. Farooq Sabir
    affiliation: University of the Punjab

date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to GExPipe}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 7,
  fig.height = 5
)
```

# Introduction

`GExPipe` is a Shiny-first workflow for bulk RNA-seq and microarray
analysis. It provides a single guided pipeline that connects data
retrieval, quality control, normalization, differential expression,
network analysis, and downstream interpretation — without requiring users
to manually stitch many separate tools together.

Within Bioconductor, `GExPipe` builds on established infrastructure such
as `r BiocStyle::Biocpkg("GEOquery")`, `r BiocStyle::Biocpkg("limma")`,
`r BiocStyle::Biocpkg("DESeq2")`, `r BiocStyle::Biocpkg("edgeR")`,
`r BiocStyle::Biocpkg("WGCNA")`, and
`r BiocStyle::Biocpkg("clusterProfiler")`. For visualization it uses
`r BiocStyle::CRANpkg("ggplot2")` and
`r BiocStyle::CRANpkg("pheatmap")`.

# Installation

## From Bioconductor (recommended once accepted)

```{r install-bioc, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# dependencies = TRUE installs all Imports + Suggests (~54 packages)
# so every app feature works out of the box.
BiocManager::install("GExPipe", dependencies = TRUE)
```

## From GitHub (current development version)

Use this block on **any system** — Windows, macOS, or Linux. It handles
DLL locks, stale lock directories, and compilation automatically:

```{r install-github, eval=FALSE}
if (!requireNamespace("shiny", quietly = TRUE))
    install.packages("shiny")

# 1. Use Windows-native HTTP — bypasses locked curl.dll on Windows
#    (harmless on macOS / Linux; wininet falls back to libcurl there)
options(download.file.method = "wininet")

# 2. Remove any 00LOCK-* directories left by a previously failed install.
#    These cause "cannot download any files" for all subsequent installs.
unlink(list.files(.libPaths()[1], pattern = "^00LOCK-", full.names = TRUE),
       recursive = TRUE, force = TRUE)

# 3. Download and run — global.R installs all missing/outdated packages
#    automatically via a background subprocess (no DLL locks, no prompts).
#    First run: ~10–30 min. Every subsequent run: seconds.
shiny::runGitHub(
  "GExPipe", "safarafique",
  ref     = "main",
  subdir  = "inst/shinyapp",
  destdir = tempfile()
)
```

> **Why these three steps?**
> On Windows, R locks every DLL of a loaded package. Step 1 switches to
> the Windows-native HTTP stack so downloads never depend on `curl.dll`.
> Step 2 clears broken lock directories from any previous failed install.
> Step 3 downloads and runs the app — `global.R` handles all package
> installation automatically in a background process that starts with
> nothing loaded and therefore has no DLL locks.

> **Rtools:** If any package requires compilation (`needs_compilation:
> TRUE`), install
> [Rtools45](https://cran.r-project.org/bin/windows/Rtools/) once.
> After that, R compiles source packages automatically with no extra
> steps.

## Alternative: `pak` (best for version-conflict resolution)

`pak` resolves the full dependency graph before downloading and
automatically upgrades any package below its minimum required version
(e.g. `rlang` too old for `dplyr`).

> **Important — `pak` also uses `libcurl` internally.**
> `options(download.file.method = "wininet")` does **not** affect `pak`.
> If `curl.dll` is locked in your current session, `pak` will fail with
> the same download error. **Restart R first** (`Ctrl+Shift+F10` in
> RStudio) to release all DLL locks, then run `pak` immediately before
> anything else loads `curl`:

```{r install-pak, eval=FALSE}
# Run AFTER restarting R (Ctrl+Shift+F10) — before anything loads curl.dll
install.packages("pak")
pak::pkg_install("safarafique/GExPipe")

# Launch:
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

If you cannot restart R (e.g. inside a running session), use the
**3-step GitHub block above** instead — it never requires a restart
because installation runs in a separate subprocess with nothing loaded.

# Launch the Shiny App

> **R version:** Use **R 4.5.x or 4.6.x stable**. Pre-release builds
> (e.g. R 4.6.0 alpha) have no pre-built binaries — install the latest
> stable from <https://cran.r-project.org/bin/windows/base/>.

## Option A — From GitHub (no install required)

Copy the block from the **Installation → From GitHub** section above.
It is safe for every user on every system and is the recommended
starting point for first-time use, classroom use, and quick tests.

What happens after you run it:

1.  `runGitHub()` downloads the latest `main` commit from GitHub into a
    temporary folder.
2.  `global.R` runs and installs all missing or outdated packages via a
    background Rscript subprocess (zero DLL locks, zero prompts).
3.  The console shows a live loading list — `✓ version` or `✗ MISSING`
    for each of the 54 packages.
4.  The app opens directly to **Step 1 — Download Data**.

## Option B — Installed package (recommended for regular use)

Install once, then launch with a single line every session:

```{r install-then-launch, eval=FALSE}
# Install (first time only)
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install(
    "safarafique/GExPipe",
    dependencies = TRUE,
    INSTALL_opts = "--no-staged-install",
    force        = TRUE
)
```

```{r launch-app, eval=FALSE}
# Every subsequent session — one line
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

What happens when you run it:

1.  `runGExPipe()` scans all 54 dependencies (under one second when
    everything is installed).
2.  Any missing or outdated package is installed in a **background
    Rscript process** — no session restart needed.
3.  The console prints a live loading list before the browser opens.
4.  `shiny::runApp()` starts the server and opens the app directly to
    **Step 1 — Download Data**.

> `runGExPipe()` follows the Bioconductor convention — it returns a
> Shiny app object; `shiny::runApp()` starts the actual server.

## Option C — Google Colab (cloud, no local R required)

Colab has no local browser, so the app must listen on all interfaces and
be exposed through a tunnel.

**Switch Colab to the R runtime:** Runtime → Change runtime type → R,
then run these cells:

**Cell 1 — Start ngrok tunnel** (free token at <https://ngrok.com>):

```{r colab-r-cell1, eval=FALSE}
NGROK_TOKEN <- "PASTE_YOUR_TOKEN_HERE"

system(paste0(
  "wget -q https://bin.equinox.io/c/bNyj1mQVY4c/",
  "ngrok-v3-stable-linux-amd64.tgz -O ngrok.tgz",
  " && tar xzf ngrok.tgz"
), wait = TRUE)
system(paste0("./ngrok config add-authtoken ", NGROK_TOKEN), wait = TRUE)
system("./ngrok http 3838 > /dev/null 2>&1 &", wait = FALSE)
Sys.sleep(3)

url <- system(paste0(
  "curl -s http://localhost:4040/api/tunnels | ",
  "python3 -c \"import sys,json; t=json.load(sys.stdin)['tunnels']; ",
  "print(t[0]['public_url'] if t else 'ngrok not ready')\""
), intern = TRUE)
cat("GExPipe URL:", url, "\n")
```

**Cell 2 — Run GExPipe** (blocks while the app is alive — normal):

```{r colab-r-cell2, eval=FALSE}
options(timeout = 3600)
if (!requireNamespace("shiny", quietly = TRUE))
  install.packages("shiny", repos = "https://cloud.r-project.org",
                   quiet = TRUE)

shiny::runGitHub(
  "GExPipe", "safarafique",
  ref            = "main",
  subdir         = "inst/shinyapp",
  destdir        = tempfile(),
  launch.browser = FALSE,
  host           = "0.0.0.0",
  port           = 3838L
)
```

# Troubleshooting

| Error | Cause | Fix |
|---|---|---|
| `download from github.com failed` | `curl.dll` locked | Use the **3-step GitHub block** (wininet + lock clean + runGitHub) |
| `pak` fails to download packages | `pak` also uses `libcurl` — same lock | Restart R (`Ctrl+Shift+F10`), then run `pak` immediately |
| `packages in use and will not be installed` | DLL locked by running session | Same **3-step block** — no restart needed |
| `namespace rlang X.Y loaded but >= Z required` | Outdated package | Restart R, then `pak::pkg_install("safarafique/GExPipe")` |
| `needs_compilation: TRUE` fails | No C compiler | Install [Rtools45](https://cran.r-project.org/bin/windows/Rtools/) |
| Corporate firewall / proxy | Network blocked | `Sys.setenv(https_proxy = "http://proxy:port")` then retry |
| No internet | Offline system | Download ZIP from GitHub, then `install.packages("GExPipe-main.zip", repos = NULL, type = "source")` |
| `lazy-load database is corrupt` | Windows staged-install bug | Reinstall with `INSTALL_opts = "--no-staged-install"` |

**Still stuck?** Restart R (`Ctrl+Shift+F10` in RStudio) and run the
3-step block again. The restart releases all DLL locks; the second run
opens the app immediately because packages are already installed.

# Advanced: Pre-install or Update Packages

`gexpipe_setup()` installs or updates all dependencies without launching
the app — useful after an R upgrade or to force a full package refresh:

```{r setup-advanced, eval=FALSE}
library(GExPipe)
gexpipe_setup(update = TRUE, optional = TRUE, launch = FALSE)

# Then launch normally
shiny::runApp(GExPipe::runGExPipe(), port = 3838L)
```

# Workflow Example (evaluated)

This section demonstrates three core tasks using bundled example data:

-   loading expression and metadata tables,
-   running basic normalization and gene intersection,
-   visualizing results (PCA and expression heatmap).

## Load package and example data

```{r load-data}
library(GExPipe)
library(ggplot2)

expr_path <- system.file("extdata", "vignette_expression.csv",
                          package = "GExPipe")
meta_path <- system.file("extdata", "vignette_sample_metadata.csv",
                          package = "GExPipe")

stopifnot(nzchar(expr_path), file.exists(expr_path))
stopifnot(nzchar(meta_path), file.exists(meta_path))

expr_tab <- read.csv(expr_path, check.names = FALSE,
                     stringsAsFactors = FALSE)
meta_tab <- read.csv(meta_path, check.names = FALSE,
                     stringsAsFactors = FALSE)

expr <- as.matrix(expr_tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- expr_tab[[1]]

rownames(meta_tab) <- meta_tab$SampleID
meta_tab <- meta_tab[colnames(expr), , drop = FALSE]

dim(expr)
head(meta_tab)
```

## Normalize and intersect datasets

`gexp_normalize_and_intersect()` accepts a named list of expression
matrices and returns a combined genes × samples matrix after
within-dataset normalization and global quantile normalization.

```{r process-data}
if (!"Dataset" %in% colnames(meta_tab))
  meta_tab$Dataset <- "D1"

datasets <- unique(meta_tab$Dataset)
micro_expr_list <- setNames(
  lapply(datasets, function(ds) {
    ids <- rownames(meta_tab)[meta_tab$Dataset == ds]
    expr[, ids, drop = FALSE]
  }),
  datasets
)

norm_res <- gexp_normalize_and_intersect(
  micro_expr_list    = micro_expr_list,
  rna_counts_list    = list(),
  micro_norm_method  = "quantile",
  rnaseq_norm_method = "TMM",
  de_method          = "limma"
)

cat("Combined matrix:", dim(norm_res$combined_expr), "\n")
cat("Common genes   :", length(norm_res$common_genes), "\n")
```

## PCA plot

```{r pca-plot}
expr_pca <- norm_res$combined_expr
keep     <- apply(expr_pca, 1, var, na.rm = TRUE) > 1e-8
pca      <- prcomp(t(expr_pca[keep, ]), center = TRUE, scale. = TRUE)

pca_df <- data.frame(
  PC1     = pca$x[, 1],
  PC2     = pca$x[, 2],
  Dataset = meta_tab[rownames(pca$x), "Dataset"],
  stringsAsFactors = FALSE
)
var_pct <- round(summary(pca)$importance[2, 1:2] * 100, 1)

ggplot(pca_df, aes(PC1, PC2, color = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title    = "PCA of bundled example data",
    subtitle = "After normalization and gene intersection",
    x = paste0("PC1 (", var_pct[1], "% variance)"),
    y = paste0("PC2 (", var_pct[2], "% variance)")
  )
```

## Expression heatmap

```{r heatmap-plot}
top_var <- head(
  order(apply(norm_res$combined_expr, 1, var, na.rm = TRUE),
        decreasing = TRUE),
  min(30L, nrow(norm_res$combined_expr))
)

if (length(top_var) > 1L) {
  pheatmap::pheatmap(
    norm_res$combined_expr[top_var, , drop = FALSE],
    scale         = "row",
    show_rownames = length(top_var) <= 30L,
    main          = "Top variable genes (bundled example)"
  )
}
```

# App Walkthrough — 15-Step Pipeline

The Shiny app is organized as a fixed 15-step pipeline. Each step must
be completed before the next becomes active.

## Step 1 — Download Data

**Purpose:** Retrieve GEO datasets and build a combined expression matrix.

**Inputs:** Platform (Microarray / RNA-seq / Merged) and one or more
GSE IDs (e.g. `GSE89076, GSE50760`).

**What the app does:** Downloads series matrix files via `GEOquery`,
maps probe or Ensembl IDs to HGNC symbols, averages duplicate probes,
and merges datasets keeping only genes present in every dataset.

**Check before continuing:** Log shows "Download complete" and sample
count matches the GEO record.

------------------------------------------------------------------------

## Step 2 — Quality Control

**Purpose:** Identify and remove outlier samples before normalization.

**What the app does:** Draws boxplots and density plots; flags
statistical outliers using Mahalanobis distance (*p* < 0.001 default).

**Check:** At least 3 samples per group must remain after removal.

------------------------------------------------------------------------

## Step 3 — Normalize

**Purpose:** Make expression values comparable across samples and
datasets.

**Methods:** Auto (quantile for microarray, TMM for RNA-seq) or Manual
(quantile, log2, VST, RLE, upper quantile).

**Check:** After-normalization boxplot shows medians at the same level.

------------------------------------------------------------------------

## Step 4 — Assign Groups

**Purpose:** Label each sample as *Normal*, *Disease*, or *Exclude*.

**Inputs:** Choose the metadata column containing condition labels, then
map each unique value to Normal / Disease / Exclude.

**Check:** Summary bar shows ≥ 3 Normal and ≥ 3 Disease samples.

------------------------------------------------------------------------

## Step 5 — Batch Correction

**Purpose:** Remove technical variation from different experiments or
sequencing runs.

**Methods:** ComBat (parametric empirical Bayes) or limma
removeBatchEffect.

**Note:** Automatically skipped when only one dataset is loaded.

**Check:** Post-correction PCA shows Normal and Disease samples
intermingled across datasets.

------------------------------------------------------------------------

## Step 6 — Differential Expression

**Purpose:** Identify genes significantly up- or down-regulated between
Normal and Disease.

**Methods:** limma (microarray), limma-voom / DESeq2 / edgeR (RNA-seq
counts). Default cutoffs: FDR < 0.05, |log₂FC| > 1.

**Outputs:** Volcano plot, heatmap, downloadable results table.

------------------------------------------------------------------------

## Step 7 — WGCNA (Co-expression Network)

**Purpose:** Group genes into co-expression modules and correlate each
module with the trait.

**Inputs:** Soft threshold β (choose where R² ≥ 0.80 on the topology
plot), minimum module size (default 30), merge cut height (default 0.25).

**Outputs:** Scale-free topology plot, dendrogram, module–trait heatmap.

**Check:** At least one module with trait correlation *p* < 0.05.

------------------------------------------------------------------------

## Step 8 — Common Genes (DEG ∩ WGCNA Hub Genes)

**Purpose:** Intersect DEGs with hub genes from trait-associated modules
to produce a high-confidence candidate gene list.

**What the app does:** Computes the intersection and runs GO / KEGG
over-representation analysis via `clusterProfiler`.

**Outputs:** Venn/UpSet diagram, GO and KEGG dot plots, downloadable
table.

------------------------------------------------------------------------

## Step 9 — PPI Network

**Purpose:** Map common genes onto a protein–protein interaction network
to identify hub proteins.

**Inputs:** STRINGdb confidence score cutoff (default 400), network layout.

**Outputs:** Interactive network plot, hub gene table with degree,
betweenness, and eigenvector centrality (downloadable).

------------------------------------------------------------------------

## Step 10 — Machine-Learning Feature Selection

**Purpose:** Identify the smallest gene set that best distinguishes
Normal from Disease.

**Methods:** LASSO, Elastic Net, Ridge, Random Forest, SVM-RFE, Boruta,
sPLS-DA, XGBoost + SHAP. Final features are genes selected by ≥ 2 algorithms.

**Outputs:** Per-algorithm importance plots, SHAP summary, final
candidate gene table.

------------------------------------------------------------------------

## Step 11 — Validation

**Purpose:** Confirm that candidate genes generalize beyond training data.

**Types:** External (independent GEO GSE ID downloaded automatically) or
Internal (70/30 stratified split).

**Check:** Validation accuracy ≥ 70%.

------------------------------------------------------------------------

## Step 12 — ROC Analysis

**Purpose:** Quantify diagnostic accuracy of each candidate gene
individually.

**What the app does:** Fits simple logistic regression per gene,
computes ROC with `pROC`, reports AUC with 95% CI.

**Check:** Retain genes with AUC ≥ 0.80 in both training and validation.

------------------------------------------------------------------------

## Step 13 — Nomogram

**Purpose:** Build a clinically interpretable scoring model from the
final gene panel.

**What the app does:**

1. Fits binary logistic regression with `rms::lrm()`.
2. Draws a nomogram (each gene has a point scale; sum gives disease
   probability).
3. Draws a calibration curve (predicted vs. observed).
4. Runs Decision Curve Analysis via `rmda`.

------------------------------------------------------------------------

## Step 14 — GSEA

**Purpose:** Identify enriched biological pathways in samples with high
vs. low expression of each final gene.

**Gene sets:** GO (BP, MF, CC), KEGG, MsigDB (Hallmark, C2, C5).

**Outputs:** Ridge plot, enrichment plot, results table with NES and
leading-edge genes.

------------------------------------------------------------------------

## Step 15 — Summary Report

**Purpose:** Generate a structured narrative and downloadable PDF
summarizing the entire analysis.

**Outputs:** Auto-generated paragraph covering dataset details, DEG
counts, WGCNA findings, common genes, PPI hubs, ML features, validation
AUCs, and top GSEA pathways. Renders to PDF via `rmarkdown`.

# App Screenshots

```{r screenshots, echo=FALSE, results='asis'}
shot_dir  <- system.file("extdata", "vignette-screenshots",
                          package = "GExPipe")
shot_meta <- list(
  list(file = "step1_download.png",  label = "Step 1 — Download Data"),
  list(file = "step2_qc.png",        label = "Step 2 — Quality Control"),
  list(file = "step6_de.png",        label = "Step 6 — Differential Expression"),
  list(file = "step9_ppi.png",       label = "Step 9 — PPI Network"),
  list(file = "step15_summary.png",  label = "Step 15 — Results Summary")
)

if (nzchar(shot_dir)) {
  for (item in shot_meta) {
    p <- file.path(shot_dir, item$file)
    if (file.exists(p)) {
      cat("\n\n## ", item$label, "\n\n", sep = "")
      print(knitr::include_graphics(p))
    }
  }
}
```

# Session Information

```{r sessionInfo}
sessionInfo()
```
