---
title: "Getting Started with GExPipe"
author: "Safa Rafique, Naeem Mehmood Ashraf, Farooq Sabir"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 3
package: GExPipe
bibliography: ../inst/references.bib
vignette: >
  %\VignetteIndexEntry{GExPipe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4
)
set.seed(123)
# Cap WGCNA parallelism during vignette builds (Bioconductor / CRAN check machines).
if (requireNamespace("WGCNA", quietly = TRUE)) {
  WGCNA::disableWGCNAThreads()
}
```

# Introduction

`GExPipe` is an **R** package and **Shiny** application for end-to-end analysis of **bulk RNA-seq** and **microarray** gene expression data. It targets GEO-archived studies and integrates quality control, normalization, batch correction, WGCNA, pathway analysis, and downstream machine learning in one workflow. Studies in the [Gene Expression Omnibus (GEO)](https://www.ncbi.nlm.nih.gov/geo/) [@edgar2002gene] are a primary use case. The interactive app guides users through download, QC, normalization, batch correction, differential expression, co-expression network analysis (WGCNA), pathway enrichment (e.g. **clusterProfiler** [@yu2012clusterprofiler]), protein–protein interaction mining, and machine learning–based biomarker exploration.

Exported **R** functions mirror the app so workflows can be scripted, tested, and reproduced outside the GUI—aligned with Bioconductor expectations for Shiny software packages [@chang2024shiny]. Differential expression uses **limma** [@ritchie2015limma] and, when appropriate, **DESeq2** [@love2014moderated] and **edgeR**. WGCNA-related steps follow standard weighted correlation network practice [@langfelder2008wgcna].

This vignette uses **bundled example matrices and sample metadata** through **WGCNA preparation**, then **downstream steps** matching the Shiny app’s **Common genes**, **GO/KEGG**, **PPI**, **machine learning**, **ROC**, **validation**, and **GSEA** tabs (`clusterProfiler`, `org.Hs.eg.db`, `STRINGdb`, `randomForest`, `pROC`, `msigdbr`, etc.). **STRING** and **KEGG REST** calls are wrapped in `tryCatch()` so builds succeed offline. For GEO download examples, see the optional section at the end of this vignette.

# Pipeline flow (all app steps)

This vignette follows the app flow end-to-end. The downstream sections are runnable during knitting, with network-dependent steps handled safely.

- **Step 1**: Parse GEO IDs and analysis settings
- **Step 2**: Quality control
- **Step 3**: Normalization
- **Step 4**: Group assignment
- **Step 5**: Batch correction
- **Step 6**: Differential expression
- **Step 7**: WGCNA preparation
- **Step 8**: Common genes (DEG intersection)
- **Step 9**: GO/KEGG enrichment and PPI
- **Step 10**: Machine learning
- **Step 11**: Validation
- **Step 12**: ROC
- **Step 13**: Nomogram-style risk modeling
- **Step 14**: GSEA (including MsigDB inputs)
- **Step 15**: Results summary

# Installation

**GExPipe** depends on core Bioconductor packages (**limma**, **DESeq2**, **clusterProfiler**, and others). Use **BiocManager** so versions stay compatible with your R installation. Once **GExPipe** is available from Bioconductor, install with:

```{r install-gexpipe-bioc-note, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("GExPipe", dependencies = TRUE)
```

For development builds, install **GExPipe** (including the Shiny app) from GitHub: [https://github.com/safarafique/GExPipe](https://github.com/safarafique/GExPipe).

```{r install-gexpipe-github, eval=FALSE}
devtools::install_github("safarafique/GExPipe")
```

# Quick Start (GEO workflow)

## Load packages

```{r library-import, message=FALSE}
library(BiocStyle)
if (requireNamespace("GExPipe", quietly = TRUE)) {
  library(GExPipe)
} else {
  if (!requireNamespace("pkgload", quietly = TRUE)) {
    stop(
      "GExPipe is not installed. Install it, or install 'pkgload' ",
      "and run from the package root to load from source."
    )
  }
  pkgload::load_all(".", quiet = TRUE, export_all = FALSE)
}
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggplot2)
library(randomForest)
library(pROC)
library(msigdbr)
library(STRINGdb)
```

# Bundled example data

Expression and phenotype tables live in `inst/extdata`. The helper below resolves paths both from an installed package and from a source checkout.

```{r read-extdata}
vignette_extdata_file <- function(name) {
  p <- system.file("extdata", name, package = "GExPipe")
  if (nzchar(p) && file.exists(p)) {
    return(p)
  }
  candidates <- c(
    file.path(getwd(), "..", "inst", "extdata", name),
    file.path(getwd(), "..", "..", "inst", "extdata", name),
    file.path(getwd(), "inst", "extdata", name)
  )
  hits <- candidates[file.exists(candidates)]
  if (length(hits) == 0L) {
    stop("Could not find bundled vignette file: ", name)
  }
  normalizePath(hits[1], winslash = "/", mustWork = TRUE)
}

expr_path <- vignette_extdata_file("vignette_expression.csv")
meta_path <- vignette_extdata_file("vignette_sample_metadata.csv")

tab <- read.csv(expr_path, check.names = FALSE, stringsAsFactors = FALSE)
meta <- read.csv(meta_path, check.names = FALSE, stringsAsFactors = FALSE)

expr <- as.matrix(tab[, -1, drop = FALSE])
storage.mode(expr) <- "numeric"
rownames(expr) <- tab[[1]]
colnames(expr) <- gsub("^X", "", colnames(expr))

rownames(meta) <- meta$SampleID
meta <- meta[colnames(expr), , drop = FALSE]

dim(expr)
head(meta)
```

# Step 1: Parse GEO study identifiers

The same identifiers you would type into Step 1 of the app can be parsed programmatically (comma-separated **GSE** IDs, platform mode, DE method).

```{r parse-gse-inputs}
parsed <- gexp_parse_gse_inputs(
  analysis_type = "merged",
  rnaseq_gses = "GSE50760",
  microarray_gses = "GSE89076",
  dataset_mode = "multi"
)
parsed$rnaseq_ids
parsed$micro_ids
```

Example used in this vignette:

- RNA-seq input: `"GSE50760"`
- Microarray input: `"GSE89076"`

For these inputs, the parsed result should include:

```{r parse-gse-expected}
stopifnot("GSE50760" %in% parsed$rnaseq_ids)
stopifnot("GSE89076" %in% parsed$micro_ids)

data.frame(
  Input = c("rnaseq_gses", "microarray_gses"),
  Entered = c("GSE50760", "GSE89076"),
  Parsed_IDs = c(
    paste(parsed$rnaseq_ids, collapse = ", "),
    paste(parsed$micro_ids, collapse = ", ")
  ),
  stringsAsFactors = FALSE
)
```

Possible outcomes at this step are:

- valid IDs are parsed and carried forward,
- empty inputs return empty ID vectors,
- malformed IDs are ignored or flagged by downstream download checks.

# Steps 2-4: QC, normalization, and group setup

For reproducible package checks, the main workflow below uses bundled example matrices from `inst/extdata` (offline-safe and deterministic).

```{r normalize-intersect}
source_note <- "Using bundled example data (check-friendly)."
cols_d1 <- rownames(meta)[meta$Dataset == "D1"]
cols_d2 <- rownames(meta)[meta$Dataset == "D2"]
micro_expr_list <- list(
  D1 = expr[, cols_d1, drop = FALSE],
  D2 = expr[, cols_d2, drop = FALSE]
)
rna_counts_list <- list()

norm <- gexp_normalize_and_intersect(
  micro_expr_list = micro_expr_list,
  rna_counts_list = rna_counts_list,
  micro_norm_method = "quantile",
  rnaseq_norm_method = "TMM",
  de_method = "limma"
)

source_note
dim(norm$combined_expr)
length(norm$common_genes)
```

```{r prepare-metadata}
um <- norm$unified_metadata

infer_condition <- function(x) {
  x <- tolower(as.character(x))
  ifelse(
    grepl("control|normal|healthy|wild.?type|wt", x),
    "Normal",
    ifelse(grepl("disease|tumou?r|cancer|case|patient|treated|knockout|ko", x), "Disease", NA_character_)
  )
}

cond_from_id <- infer_condition(um$SampleID)
if (all(is.na(cond_from_id))) {
  # Deterministic fallback to keep runnable flow when labels are absent.
  ds <- unique(um$Dataset)
  cond_from_id <- ifelse(um$Dataset == ds[1], "Normal", "Disease")
}

um$Condition <- cond_from_id
ctab <- table(um$Dataset, um$Condition)
# Auto-balance labels for demonstration if Dataset and Condition are perfectly confounded.
if (all(rowSums(ctab > 0) <= 1) || all(colSums(ctab > 0) <= 1)) {
  for (ds in unique(um$Dataset)) {
    idx <- which(um$Dataset == ds)
    if (length(idx) >= 2L) {
      half <- floor(length(idx) / 2)
      um$Condition[idx] <- "Disease"
      um$Condition[idx][seq_len(half)] <- "Normal"
    }
  }
  source_note <- paste0(source_note, " Auto-balanced Condition labels for demonstration outputs.")
}
table(um$Dataset, um$Condition)
```

In small bundled examples, `Dataset` and `Condition` can be strongly aligned (confounded). When that happens, downstream DE/ML models may return conservative or empty outputs (this is expected for check-safe data).

Optional demonstration remapping (manual, not run during checks):

```{r optional-remap-condition, eval=FALSE}
# If you want richer DE/ML outputs in a local demo, remap a few samples
# so each dataset contains both Normal and Disease labels.
#
# Example strategy:
# flip_idx <- which(um$Dataset == "D1")[1:2]
# um$Condition[flip_idx] <- "Disease"
# flip_idx2 <- which(um$Dataset == "D2")[1:2]
# um$Condition[flip_idx2] <- "Normal"
#
# table(um$Dataset, um$Condition)
```

## Step 2 QC checks (helper outputs)

```{r qc-helpers}
qc_out <- gexp_qc_detect_outliers(norm$combined_expr, top_n = 100)
length(qc_out$all_outliers)

qc_summary <- gexp_qc_gene_overlap_summary(
  all_genes_list = list(
    D1 = rownames(micro_expr_list$D1),
    D2 = rownames(micro_expr_list$D2)
  ),
  common_genes = norm$common_genes
)
qc_summary
```

# Step 5: Batch correction

After integration, batch effects (e.g. **Dataset**) can still dominate variation. **Batch correction** adjusts expression while retaining biological signal; you should verify that samples cluster more by **Condition** (or phenotype) than by batch.

```{r batch-correct}
batch_res <- gexp_batch_correct(
  expr = norm$combined_expr,
  metadata = um[, c("Dataset", "Condition")],
  variance_percentile = 25,
  method = "limma"
)

dim(batch_res$expr_filtered)
dim(batch_res$batch_corrected)
```

## PCA visualization (post-batch correction)

A quick **PCA** on batch-corrected values helps confirm that correction worked as intended (e.g. **Condition**-driven separation visible; **Dataset** less dominant than before correction in exploratory plots).

```{r batch-correct-pca, fig.cap = "PCA of samples after batch correction (bundled example data)."}
sx <- colnames(batch_res$batch_corrected)
bc <- batch_res$batch_corrected
pca_fit <- prcomp(t(bc), center = TRUE, scale. = TRUE)
pca_df <- data.frame(
  PC1 = pca_fit$x[sx, 1L],
  PC2 = pca_fit$x[sx, 2L],
  Dataset = um[sx, "Dataset", drop = TRUE],
  Condition = um[sx, "Condition", drop = TRUE],
  row.names = sx,
  stringsAsFactors = FALSE
)

ggplot(pca_df, aes(x = PC1, y = PC2, color = Condition, shape = Dataset)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(
    title = "PCA: post-batch correction",
    subtitle = "Samples should cluster more by Condition than by Dataset"
  )
```

# Step 6: Differential expression

```{r run-de}
de_res <- gexp_run_de(
  expr = batch_res$batch_corrected,
  metadata = um[, c("Dataset", "Condition")],
  method = "limma",
  logfc_cutoff = 0.25,
  padj_cutoff = 0.10
)

de_df <- de_res$de_results
table(de_df$Significance)
head(de_df[, c("Gene", "logFC", "adj.P.Val", "Significance")], 8)
```

# Step 7: WGCNA preparation

```{r wgcna-prepare}
wgcna_prep <- gexp_wgcna_prepare(
  expr = batch_res$batch_corrected,
  metadata = um,
  gene_mode = "top_variable",
  top_genes = min(100L, nrow(batch_res$batch_corrected)),
  min_samples_frac = 0.5
)

dim(wgcna_prep$datExpr)
head(wgcna_prep$gene_variance_table, 3)
```

# Step 8: Common genes (DEG and WGCNA overlap)

In the app (**Step 8**), **common genes** are the intersection of **significant DEGs** (differential expression) and genes that fall in **trait-related WGCNA modules**. This vignette uses the same idea in compact form: significant DEGs ∩ genes kept in `gexp_wgcna_prepare()` output (`wgcna_prep$datExpr`). We also summarize overlap with **`gexp_qc_gene_overlap_summary()`** (exported helper used in QC logic).

```{r common-genes}
deg_genes <- unique(de_df$Gene[de_df$Significance != "Not Significant"])
# gexp_wgcna_prepare() returns datExpr as samples x genes, so gene names are columns.
wgcna_genes <- colnames(wgcna_prep$datExpr)
common_genes <- intersect(deg_genes, wgcna_genes)

if (length(common_genes) < 5L) {
  deg_genes <- head(de_df$Gene[order(de_df$adj.P.Val)], 40)
  common_genes <- intersect(deg_genes, wgcna_genes)
}

length(common_genes)
head(common_genes, 10)

gexp_qc_gene_overlap_summary(
  all_genes_list = list(DEG = deg_genes, WGCNA = wgcna_genes),
  common_genes = common_genes
)
```

The app also offers **Venn** and **UpSet** views for the same lists; this vignette sticks to **tables** only.

```{r common-genes-upset-check}
up_meta <- gexp_qc_prepare_upset_data(list(DEG = deg_genes, WGCNA = wgcna_genes))
up_meta$ok ## TRUE when both lists are non-empty (data ready for UpSet in the app)
```

# Step 9: Gene Ontology enrichment (BP, MF, CC)

We map **gene symbols** to **Entrez IDs** with `clusterProfiler::bitr()`, then run **`enrichGO()`** separately for **Biological Process**, **Molecular Function**, and **Cellular Component**—the same ORA pattern as the app’s GO panels on the **common-genes** path.

```{r go-map-ids}
eg <- tryCatch(
  clusterProfiler::bitr(
    common_genes,
    fromType = "SYMBOL",
    toType = "ENTREZID",
    OrgDb = org.Hs.eg.db::org.Hs.eg.db
  ),
  error = function(e) data.frame(SYMBOL = character(), ENTREZID = character())
)
nrow(eg)
```

```{r go-enrichment-all}
go_params <- list(
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  pAdjustMethod = "BH",
  pvalueCutoff = 0.2,
  qvalueCutoff = 0.3,
  readable = TRUE
)

ego_list <- if (nrow(eg) == 0L) {
  list(BP = NULL, MF = NULL, CC = NULL)
} else {
  lapply(c(BP = "BP", MF = "MF", CC = "CC"), function(ont) {
    tryCatch(
      do.call(
        clusterProfiler::enrichGO,
        c(list(gene = eg$ENTREZID, ont = ont), go_params)
      ),
      error = function(e) NULL
    )
  })
}

vapply(ego_list, function(x) if (is.null(x)) 0L else nrow(as.data.frame(x)), integer(1))
```

```{r go-tables}
for (nm in names(ego_list)) {
  ed <- if (is.null(ego_list[[nm]])) data.frame() else as.data.frame(ego_list[[nm]])
  cat("## ", nm, " — n terms:", nrow(ed), "\n", sep = "")
  if (nrow(ed) > 0L) {
    show_cols <- intersect(c("Description", "pvalue", "p.adjust", "Count"), names(ed))
    print(head(ed[, show_cols, drop = FALSE], 3))
  }
}
```

# Step 9: KEGG pathway enrichment (over-representation)

`enrichKEGG()` uses the **KEGG REST API** (needs **network** access for live queries). The chunk always runs; if you are offline or the service is unreachable, the result is empty.

```{r kegg-enrichment}
ekegg <- if (nrow(eg) == 0L) {
  NULL
} else {
  tryCatch(
    clusterProfiler::enrichKEGG(
      gene = eg$ENTREZID,
      organism = "hsa",
      pvalueCutoff = 0.2,
      qvalueCutoff = 0.3
    ),
    error = function(e) NULL
  )
}

kegg_df <- if (!is.null(ekegg)) as.data.frame(ekegg) else data.frame()
nrow(kegg_df)
if (nrow(kegg_df) > 0L) {
  head(kegg_df[, intersect(c("ID", "Description", "p.adjust", "Count"), names(kegg_df))], 5)
}
```

# Step 9: Protein-protein interactions (STRING)

The **PPI** tab maps your gene list to **STRING** identifiers and retrieves **experimental / database** edges above a score threshold. Below we cap the query size for speed; increase `head(common_genes, …)` for a fuller network when online.

```{r ppi-string}
ppi_res <- tryCatch(
  {
    if (length(common_genes) < 3L) {
      stop("Need at least three common genes for STRING.")
    }
    string_db <- STRINGdb$new(version = "12", species = 9606, score_threshold = 400)
    mapped <- string_db$map(
      data.frame(gene = head(common_genes, 120L)),
      "gene",
      removeUnmappedRows = TRUE
    )
    if (nrow(mapped) < 2L) {
      stop("Too few STRING mappings for this gene list.")
    }
    hits <- string_db$get_interactions(mapped$STRING_id)
    list(
      n_edges = nrow(hits),
      msg = paste0("Retrieved ", nrow(hits), " STRING interaction rows (score ≥ 400).")
    )
  },
  error = function(e) {
    list(n_edges = NA_integer_, msg = paste0("PPI skipped: ", conditionMessage(e)))
  }
)

ppi_res$msg
ppi_res$n_edges
```

# Step 10: Machine learning (supervised classification)

**Step 10 (ML)** in the app trains classifiers on expression features (often **common genes**). Here we fit a **random forest** on **batch-corrected** values, **Normal vs Disease**, and show a short **importance** table when the fit succeeds.

```{r ml-rf}
genes_ml <- common_genes[seq_len(min(25L, length(common_genes)))]
sid <- colnames(batch_res$batch_corrected)
X <- t(batch_res$batch_corrected[genes_ml, sid, drop = FALSE])
y <- factor(um[sid, "Condition"], levels = c("Normal", "Disease"))

X[!is.finite(X)] <- NA
keep <- stats::complete.cases(X) & !is.na(y)
X <- X[keep, , drop = FALSE]
y <- droplevels(y[keep])

if (ncol(X) > 0L) {
  vr <- apply(X, 2L, function(z) stats::var(z, na.rm = TRUE))
  okc <- is.finite(vr) & vr > 0
  X <- if (any(okc)) X[, okc, drop = FALSE] else X[, integer(), drop = FALSE]
}

rf_fit <- if (isTRUE(nrow(X) >= 3L) && ncol(X) >= 1L && length(unique(y)) >= 2L) {
  tryCatch(
    randomForest::randomForest(x = X, y = y, ntree = 150L, importance = TRUE),
    error = function(e) NULL
  )
} else {
  NULL
}
rf_fit

if (!is.null(rf_fit) && !is.null(rf_fit$importance) && nrow(rf_fit$importance) > 0L) {
  imp <- as.data.frame(rf_fit$importance)
  imp$Gene <- rownames(imp)
  imp <- imp[order(imp$MeanDecreaseGini, decreasing = TRUE), , drop = FALSE]
  head(imp[, c("Gene", "MeanDecreaseGini"), drop = FALSE], min(12L, nrow(imp)))
}
```

# Step 11: Validation (external-style holdout)

**Step 11 (Validation)** in the app supports **external cohorts**. Here we **simulate** that by treating **Dataset D1** as **training** and **Dataset D2** as **external test**—same genes (`common_genes` subset), shared **Condition** labels—so the code path is fully **runnable** on bundled data.

```{r validation-external-style}
val_auc <- NA_real_
val_txt <- "Validation not run."

genes_val <- common_genes[seq_len(min(20L, length(common_genes)))]
sid_tr <- intersect(rownames(um)[um$Dataset == "D1"], colnames(batch_res$batch_corrected))
sid_te <- intersect(rownames(um)[um$Dataset == "D2"], colnames(batch_res$batch_corrected))

if (length(genes_val) >= 5L && length(sid_tr) >= 3L && length(sid_te) >= 2L) {
  Xtr <- t(batch_res$batch_corrected[genes_val, sid_tr, drop = FALSE])
  ytr <- droplevels(factor(um[sid_tr, "Condition"]))
  Xte <- t(batch_res$batch_corrected[genes_val, sid_te, drop = FALSE])
  yte <- factor(um[sid_te, "Condition"], levels = levels(ytr))

  keep_tr <- stats::complete.cases(Xtr) & !is.na(ytr)
  Xtr <- Xtr[keep_tr, , drop = FALSE]
  ytr <- droplevels(ytr[keep_tr])

  if (ncol(Xtr) > 0L) {
    vr <- apply(Xtr, 2L, function(z) stats::var(z, na.rm = TRUE))
    ok <- is.finite(vr) & vr > 0
    Xtr <- Xtr[, ok, drop = FALSE]
    Xte <- Xte[, colnames(Xtr), drop = FALSE]
  }

  keep_te <- stats::complete.cases(Xte) & !is.na(yte)
  Xte <- Xte[keep_te, , drop = FALSE]
  yte <- droplevels(yte[keep_te])

  if (ncol(Xtr) >= 1L && nrow(Xtr) >= 3L && length(unique(ytr)) >= 2L && nrow(Xte) >= 1L) {
    fit_ext <- tryCatch(
      randomForest::randomForest(x = Xtr, y = ytr, ntree = 150L),
      error = function(e) NULL
    )
    if (!is.null(fit_ext)) {
      pr <- tryCatch(predict(fit_ext, Xte, type = "prob"), error = function(e) NULL)
      if (is.matrix(pr) && "Disease" %in% colnames(pr) && length(unique(yte)) >= 2L) {
        roc_te <- tryCatch(
          pROC::roc(yte, pr[, "Disease"], quiet = TRUE, levels = levels(ytr)),
          error = function(e) NULL
        )
        if (!is.null(roc_te)) {
          val_auc <- as.numeric(pROC::auc(roc_te))
          val_txt <- sprintf(
            "Held-out dataset D2: AUC = %.3f (train D1 / test D2, %d genes).",
            val_auc,
            ncol(Xtr)
          )
        }
      }
    }
  }
}

val_txt
```

# Step 12: ROC curve (discrimination)

**Step 12 (ROC)** uses predicted probabilities vs true labels. We use **training-set** probabilities from the forest model as an illustrative example. For unbiased performance, prefer cross-validation or a held-out dataset.

```{r roc-demo}
roc_auc_txt <- "ROC / AUC not computed."
if (!is.null(rf_fit)) {
  oob <- tryCatch(
    predict(rf_fit, type = "prob"),
    error = function(e) NULL
  )
  if (!is.null(oob) && is.matrix(oob) && ncol(oob) >= 2L && "Disease" %in% colnames(oob)) {
    roc_obj <- tryCatch(
      pROC::roc(response = y, predictor = oob[, "Disease"], quiet = TRUE),
      error = function(e) NULL
    )
    if (!is.null(roc_obj)) {
      roc_auc_txt <- sprintf(
        "AUC (training predictions, illustrative) = %.4f",
        as.numeric(pROC::auc(roc_obj))
      )
    }
  }
}
roc_auc_txt
```

# Step 13: Nomogram-style risk model (lightweight)

The app's nomogram tab is designed for clinical risk presentation. In this vignette, we provide a lightweight, fully runnable proxy using logistic regression to estimate disease risk from top genes.

```{r nomogram-proxy}
nomo_txt <- "Nomogram-style model not run."
nomo_df <- data.frame()

if (exists("imp") && is.data.frame(imp) && nrow(imp) > 0L) {
  topg <- head(imp$Gene, min(6L, nrow(imp)))
  sx <- colnames(batch_res$batch_corrected)
  Xn <- t(batch_res$batch_corrected[topg, sx, drop = FALSE])
  yn <- factor(um[sx, "Condition"], levels = c("Normal", "Disease"))
  keep_n <- stats::complete.cases(Xn) & !is.na(yn)
  Xn <- Xn[keep_n, , drop = FALSE]
  yn <- droplevels(yn[keep_n])
  if (nrow(Xn) >= 10L && ncol(Xn) >= 2L && length(unique(yn)) == 2L) {
    dfn <- data.frame(y = yn, Xn, check.names = FALSE)
    fit_n <- tryCatch(stats::glm(y ~ ., data = dfn, family = stats::binomial()), error = function(e) NULL)
    if (!is.null(fit_n)) {
      pr <- tryCatch(stats::predict(fit_n, newdata = dfn, type = "response"), error = function(e) NULL)
      if (!is.null(pr)) {
        nomo_txt <- sprintf(
          "Nomogram-style proxy fit succeeded on %d samples and %d genes.",
          nrow(dfn), ncol(dfn) - 1L
        )
      }
      cf <- stats::coef(fit_n)
      nomo_df <- data.frame(Term = names(cf), Coefficient = unname(cf), row.names = NULL)
    }
  }
}
nomo_txt
head(nomo_df, 8)
```

# Step 14: GSEA (gene-set enrichment analysis)

**Step 14 (GSEA)** ranks **all genes** by a differential statistic (here **limma logFC**) and tests whether curated sets are enriched at the **top or bottom** of the ranking (`clusterProfiler::gseGO()`). We show **GO BP**; the app can combine this with **MsigDB** collections.

```{r gsea-gsego}
gene_list <- de_df$logFC
names(gene_list) <- de_df$Gene
gene_list <- sort(gene_list, decreasing = TRUE)

gse_bp <- tryCatch(
  clusterProfiler::gseGO(
    geneList = gene_list,
    OrgDb = org.Hs.eg.db::org.Hs.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    minGSSize = 10,
    maxGSSize = 500,
    pvalueCutoff = 0.5,
    eps = 0
  ),
  error = function(e) NULL
)

gse_df <- if (!is.null(gse_bp)) as.data.frame(gse_bp) else data.frame()
nrow(gse_df)
if (nrow(gse_df) > 0L) {
  head(gse_df[, intersect(c("Description", "NES", "pvalue", "p.adjust"), names(gse_df))], 5)
}
```

```{r gsea-kegg-optional}
## KEGG GSEA expects Entrez-ranked scores; map symbols from the same ranked list as gseGO.
gse_kegg <- NULL
gse_k_df <- data.frame()
if (length(gene_list) >= 10L) {
  kegg_map <- tryCatch(
    clusterProfiler::bitr(
      names(gene_list),
      fromType = "SYMBOL",
      toType = "ENTREZID",
      OrgDb = org.Hs.eg.db::org.Hs.eg.db
    ),
    error = function(e) data.frame()
  )
  if (nrow(kegg_map) > 5L) {
    sym_scores <- gene_list[kegg_map$SYMBOL]
    ez <- kegg_map$ENTREZID
    okm <- !is.na(sym_scores) & !is.na(ez)
    gl_entrez <- tapply(sym_scores[okm], ez[okm], mean, simplify = TRUE)
    gl_entrez <- sort(gl_entrez, decreasing = TRUE)
    gse_kegg <- tryCatch(
      clusterProfiler::gseKEGG(
        geneList = gl_entrez,
        organism = "hsa",
        minGSSize = 10,
        maxGSSize = 500,
        pvalueCutoff = 0.5,
        eps = 0
      ),
      error = function(e) NULL
    )
    gse_k_df <- if (!is.null(gse_kegg)) as.data.frame(gse_kegg) else data.frame()
  }
}
nrow(gse_k_df)
if (nrow(gse_k_df) > 0L) {
  head(
    gse_k_df[, intersect(c("ID", "Description", "NES", "p.adjust"), names(gse_k_df))],
    min(8L, nrow(gse_k_df))
  )
}
```

# Step 14: MsigDB / Hallmark gene sets (GSEA inputs)

**Step 14** can also use **Molecular Signatures Database** collections. The **`msigdbr`** package supplies tidy tables (no extra registration). Below we load **Hallmark (H)** gene sets for **Homo sapiens** as a typical ranked-list GSEA input.

```{r msigdbr-hallmark}
hmk <- msigdbr(species = "Homo sapiens", category = "H")
n_sets <- length(unique(hmk$gs_name))
n_sets
head(hmk[, c("gs_name", "gene_symbol", "entrez_gene")], 10)
```

# Step 15: Results summary

This section summarizes key downstream outputs in one place, similar to the app's final summary tab.

```{r final-summary}
summary_tbl <- data.frame(
  Metric = c(
    "Common genes (Step 8)",
    "PPI edges (Step 9)",
    "Validation AUC (Step 11)",
    "ROC AUC, illustrative (Step 12)",
    "GO GSEA terms (Step 14)",
    "KEGG GSEA terms (Step 14)"
  ),
  Value = c(
    as.character(length(common_genes)),
    as.character(ppi_res$n_edges),
    ifelse(is.na(val_auc), "NA", sprintf("%.3f", val_auc)),
    roc_auc_txt,
    as.character(nrow(gse_df)),
    as.character(nrow(gse_k_df))
  ),
  row.names = NULL
)
summary_tbl
```

## Real run output (auto-generated)

The table below is generated from the current vignette run (not hardcoded), so values reflect your real execution environment and data availability.
The bundled vignette example is designed for reproducible, check-safe execution.
For biologically meaningful enrichment terms, richer networks, and publication-quality plots, run `GExPipe` on your own real datasets (GEO IDs or local expression/metadata files).

```{r real-run-output}
run_output <- data.frame(
  Item = c(
    "Data source",
    "Expression matrix after normalization",
    "Common genes retained",
    "Detected outliers",
    "Batch-corrected matrix",
    "DE result rows",
    "Validation status",
    "GSEA GO terms",
    "GSEA KEGG terms",
    "PPI status"
  ),
  Output = c(
    if (exists("source_note")) source_note else "Bundled example data",
    if (exists("norm")) paste(dim(norm$combined_expr), collapse = " x ") else "NA",
    if (exists("common_genes")) as.character(length(common_genes)) else "NA",
    if (exists("qc_out")) as.character(length(qc_out$all_outliers)) else "NA",
    if (exists("batch_res")) paste(dim(batch_res$batch_corrected), collapse = " x ") else "NA",
    if (exists("de_df")) as.character(nrow(de_df)) else "NA",
    if (exists("val_auc") && !is.na(val_auc)) sprintf("AUC = %.3f", val_auc) else "Not available",
    if (exists("gse_df")) as.character(nrow(gse_df)) else "NA",
    if (exists("gse_k_df")) as.character(nrow(gse_k_df)) else "NA",
    if (exists("ppi_res")) as.character(ppi_res$msg) else "NA"
  ),
  stringsAsFactors = FALSE
)
run_output
```

# Interactive Shiny application

`runGExPipe()` returns a **Shiny app object** (the package does not call `shiny::runApp()` for you). Pass the result to `shiny::runApp()` in an interactive session. Use a **fixed port** (e.g. `3838`) or `launch.browser = TRUE` so the browser URL is valid; the returned object is suitable for the RStudio Viewer or an external browser.

```{r app-object}
## Faster vignette check: defer heavy namespace attach (same pattern as other vignettes).
options(
  shiny.testmode = TRUE,
  gexpipe.minimal_attach_in_testmode = TRUE,
  gexpipe.wgcna_threads = 0L
)
port_test <- httpuv::randomPort()
app <- runGExPipe(launch.browser = FALSE, port = port_test)
inherits(app, "shiny.appobj")
```

```{r launch-interactive, eval=FALSE}
# Returns the app object for the browser or RStudio Viewer.
if (interactive()) {
  app <- GExPipe::runGExPipe(launch.browser = TRUE, port = 3838L)
  shiny::runApp(app, port = 3838L)
}
```

# Further reading

- `vignettes/gexpipe-shiny-testing-ci.Rmd` — **shinytest2** smoke tests and CI notes.

# Optional GEO workflow (network-dependent)

Not evaluated during package checks; run locally when **GEOquery** access is available.

```{r optional-geo, eval=FALSE}
# Real GEO example used in app-style parsing:
work_dir <- getwd()
gexp_prepare_download_dirs(work_dir, has_micro = TRUE, has_rna = TRUE)

micro_dir <- file.path(work_dir, "micro_data")
rna_dir <- file.path(work_dir, "rna_data")

micro_dl <- gexp_download_one_microarray_gse("GSE89076", micro_dir)
rna_dl <- gexp_download_one_rnaseq_gse("GSE50760", rna_dir)

micro_dl$ok
rna_dl$ok
```

# Session Info

```{r sessioninfo}
sessionInfo()
```

# References
