---
title: "OmniVerse: tutorial and example data"
author: "Safa Rafique"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{OmniVerse: tutorial and example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

**OmniVerse** is a Shiny application for high-throughput genomic analysis of bulk RNA-seq and microarray expression data. It provides an end-to-end, interactive workflow including download/QC, normalization, differential expression, WGCNA, pathway enrichment, PPI, and ML.

This vignette shows:

- how to **run** the app,
- how to use a **small example workspace** bundled with the package (offline; no GEO download required),
- how to **save / load** a workspace for reproducible sessions.

## Run the app

```{r run-app, eval=TRUE}
library(OmniVerse)

if (interactive()) {
  runOmniVerse()
}
```

## Inputs OmniVerse can use

OmniVerse can start from:

- a **GEO download** (GSE/GPL; requires internet), or
- a **saved workspace** (`.rds`) created by OmniVerse (recommended for sharing and reproducibility).

This vignette focuses on the second option because it is deterministic and works offline.

## Example data included with the package

The package includes a small example workspace file:

```{r example-path}
ex_path <- system.file("extdata", "omniverse_example_workspace.rds", package = "OmniVerse")
ex_path
stopifnot(nzchar(ex_path), file.exists(ex_path))
```

You can inspect what is inside the workspace:

```{r inspect-workspace}
ws <- readRDS(ex_path)
names(ws)

# Key objects typically present in a workspace
if (!is.null(ws$combined_expr_raw)) dim(ws$combined_expr_raw)
if (!is.null(ws$unified_metadata)) head(ws$unified_metadata)
```

### Use the example workspace in the app (recommended)

1. Start the app with `runOmniVerse()`.
2. Go to **Step 1 (Download)**.
3. In the **workspace load** panel, choose **“Upload a .rds file”** and select the example file at:

`r ex_path`

4. Click **“Load uploaded file”**.

OmniVerse will jump to the saved step (this example opens at **QC**) and you can continue through the pipeline.

## Saving a workspace (for reproducibility)

At any point you can save your current analysis state:

- **Download workspace**: saves a `.rds` file via the browser download button.
- **Save to folder**: writes into a `saved_workspaces/` folder in the app working directory.

To resume later (or share with a collaborator), open OmniVerse and load the `.rds` from Step 1.

## Package versions

For a tested-with snapshot of important package versions, see:

```{r versions-file}
system.file("pkg_versions.txt", package = "OmniVerse")
```

## Session info

```{r session-info}
sessionInfo()
```

